---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

<!-- badges: start -->
[![R-CMD-check](https://github.com/RConsortium/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/RConsortium/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/RConsortium/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/RConsortium/OOP-WG?branch=master)
<!-- badges: end -->

## Workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

## Example

```{r error = TRUE}
library(R7)

range <- class_new("range",
  constructor = function(start, end) {
    object_new(start = start, end = end)
  },
  validator = function(x) {
    if (property(x, "end") < property(x, "start")) {
      "`end` must be greater than or equal to `start`"
    }
  },
  properties = list(
    start = "numeric",
    end = "numeric",
    property_new(
      name = "length",
      accessor = function(x) x@end - x@start
    )
  )
)

x <- range(start = 1, end = 10)

x@start

x@end

# assigning properties verifies the class
x@end <- "foo"

# assigning properties runs the validator
x@end <- 0

object_class(x)
```

## Performance

The dispatch performance should be roughly on par with S3 and S4, though as this is implemented in the package there is some overhead due to `.Call` vs `.Primitive`.

Dispatch uses a table stored in the `methods` property of the generic. This table is a nested set of hashed environments based on the classes of the methods. e.g.

`method(foo, c("character", "numeric"))` method would be stored at

`foo@methods[["character"]][["numeric"]]`

At each level the search iteratively searches up the class vector for the object.

I am not sure caching is really needed, most objects have very short class lists and the lookup is fast as each level should have relatively few items and we are using hashed environments.

```{r performance, cache = FALSE}
text <- class_new("text", parent = "character", constructor = function(text) object_new(.data = text))
number <- class_new("number", parent = "numeric", constructor = function(x) object_new(.data = x))

x <- text("hi")
y <- number(1)

foo_r7 <- generic_new(name = "foo_r7", signature = "x")
method(foo_r7, "text") <- function(x) paste0(x, "-foo")

foo_s3 <- function(x) {
  UseMethod("foo_s3")
}

foo_s3.text <- function(x) {
  paste0(x, "-foo")
}

library(methods)
setOldClass(c("number", "numeric", "r7_object"))
setOldClass(c("text", "character", "r7_object"))

setGeneric("foo_s4", function(x) standardGeneric("foo_s4"))
setMethod("foo_s4", c("text"), function(x) paste0(x, "-foo"))

# Measure performance of single dispatch
bench::mark(foo_r7(x), foo_s3(x), foo_s4(x))


bar_r7 <- generic_new("bar_r7", c("x", "y"))
method(bar_r7, list("text", "number")) <- function(x, y) paste0(x, "-", y, "-bar")

setGeneric("bar_s4", function(x, y) standardGeneric("bar_s4"))
setMethod("bar_s4", c("text", "number"), function(x, y) paste0(x, "-", y, "-bar"))

# Measure performance of double dispatch
bench::mark(bar_r7(x, y), bar_s4(x, y))
```

## TODO

- Objects
  - [x] - A class object attribute, a reference to the class object, and retrieved with `object_class()`.
  - [x] - For S3 compatibility, a class attribute, a character vector of class names.
  - [x] - Additional attributes storing properties defined by the class, accessible with `@/property()`.
- Classes
  - [x] - R7 classes are first class objects with the following
    - [x] - `name`, a human-meaningful descriptor for the class.
    - [x] - `parent`, the class object of the parent class.
    - [x] - A constructor, an user-facing function used to create new objects of this class. It always ends with a call to `object_new()` to initialize the class.
    - [x] - A validator, a function that takes the object and returns NULL if the object is valid, otherwise a character vector of error messages.
    - [x] - properties, a list of property objects
- Initialization
  - [x] - The constructor uses `object_new()` to initialize a new object, this
    - [x] - Inspects the enclosing scope to find the "current" class.
    - [ ] - Creates the prototype, by either by calling the parent constructor or by creating a base type and adding class and `object_class()` attributes to it.
    - [x] - Validates properties then adds to prototype.
    - [x] - Validates the complete object.
- Shortcuts
  - [ ] - any argument that takes a class object can instead take the name of a class object as a string
  - [x] - instead of providing a list of property objects, you can instead provide a named character vector.
- Validation
  - [x] - valid_eventually
  - [x] - valid_implicitly
- Unions
  - [ ] - Used in properties to allow a property to be one of a set of classes
  - [x] - In method dispatch as a convenience for defining a method for multiple classes
- Properties
  - [x] - Accessed using `property()` / `property<-`
  - [x] - Accessed using `@` / `@<-`
  - [x] - A name, used to label output
  - [ ] - A optional class or union
  - [x] - An optional accessor function
  - [ ] - Properties are created with `prop_new()`
- Generics
  - [x] - It knows its name and the names of the arguments in its signature
  - [x] - Calling `generic_new()` defines a new generic
  - [ ] - By convention, any argument that takes a generic function, can instead take the name of a generic function supplied as a string
- Methods
  - Registration
    - [x] - Methods are defined by calling method<-(generic, signature, method):
    - [x] - generic is a generic function.
    - [x] - signature is a
      - [x] - single class object
      - [x] - a class union
      - [x] - list of class objects/unions
      - [x] - a character vector.
    - [ ] - method is a compatible function
    - [ ] - method<- is designed to work at run-time
  - Dispatch
    - [x] - Dispatch is nested, meaning that if there are multiple arguments in the generic signature, it will dispatch on the first argument, then the second.
    - [x] - A `plot()` generic dispatching on `x`, e.g. ` plot <- function(x) { method(plot, object_class(x))(x) }`
    - [x] - A `publish()` that publishes an object `x` to a destination `y`, dispatching on both arguments, e.g. `publish <- function(x, y, ...) { method(publish, list(object_class(x), object_class(y)))(x, y, ...) }`
    - [x] - `...` is not used for dispatch
    - [x] - R7 generics can dispatch with base type objects
    - [x] - R7 generics can dispatch with S3 objects
    - [x] - R7 generics can dispatch with S4 objects
- Compatibility
  - S3
    - [x] - Since the class attribute has the same semantics as S3, S3 dispatch should be fully compatible.
    - [x] - The new generics should also be able to handle legacy S3 objects.
    - [x] - `method()` falls back to single argument S3 dispatch if the R7 dispatch fails.
    - [ ] - `method()` uses S3 group generics as well
  - S4
    - [x] - Since the new generics will fallback to S3 dispatch, they should support S4 objects just as S3 generics support them now.
- Documentation
  - [ ] - Generate index pages that list the methods for a generic or the methods with a particular class in their signature

## Questions
- What should happen if you call `method_new()` on a S3 generic?
  1. Create a new R7 generic out of the S3 generic?
  2. Just register the R7 object using registerS3method?
- Best way to support `subsutitue()` calls in methods? We need to evaluate the argument promises to do the dispatch, but we want to pass the un-evaluated promise to the call?
- on-load method registration
- how will `method_next()` work with nested multiple dispatch? How do we know which argument you want the next method for?
- If a type has only properties, what is the base type? R7 currently using VECSXP, S4 uses S4SXP
- `method_new()` vs `method()<-`, the latter while nice has drawbacks
  - can't use `method("foo")<-`
  - can't use `method(otherpkg::foo)<-`
