---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

These ideas have been implemented in the R7 package, hosted in this repository.

<!-- badges: start -->
[![R-CMD-check](https://github.com/RConsortium/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/RConsortium/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/RConsortium/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/RConsortium/OOP-WG?branch=master)
<!-- badges: end -->


## Classes and objects

```{r error = TRUE}
library(R7)

range <- new_class("range",
  constructor = function(start, end) {
    new_object(start = start, end = end)
  },
  validator = function(x) {
    if (x@end < x@start) {
      "<range>@end must be greater than or equal to <range>@start"
    }
  },
  properties = list(
    start = "numeric",
    end = "numeric",
    new_property(
      name = "length",
      class = "numeric",
      getter = function(x) x@end - x@start,
      setter = function(x, value) {
        x@end <- x@start + value
        x
      }
    )
  )
)

x <- range(start = 1, end = 10)

x@start

x@end

x@length

x@length <- 5

x@length

# incorrect properties throws an error
x@middle

# assigning properties verifies the class matches the class of the value
x@end <- "foo"

# assigning properties runs the validator
x@end <- 0

# Print methods for both R7_class objects
object_class(x)

# As well as normal R7_objects
x

# Use `.data` to refer to and retrieve the base data type, properties are
# automatically removed, but non-property attributes (such as names) are retained.

text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

y <- text(c(foo = "bar"))

names(y@.data)
```

## Generics and methods

```{r}
text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

foo <- new_generic("foo", signature = "x")

method(foo, "text") <- function(x, ...) paste0("foo-", x)

foo(text("hi"))
```

## Multiple dispatch

Multiple dispatch uses a table stored in the `methods` property of the generic.
This table is a nested set of hashed environments based on the classes of the methods. e.g.

For `method(foo, c("character", "numeric"))` the method would be stored at `foo@methods[["character"]][["numeric"]]`.

At each level the search iteratively searches along objects class vector.

```{r}
number <- new_class("number", parent = "numeric", constructor = function(x) new_object(.data = x))

bar <- new_generic("bar", signature = c("x", "y"))

method(bar, list("character", "numeric")) <- function(x, y, ...) paste0("foo-", x, ":", y)

bar(text("hi"), number(42))
```

## Calling the next method

`next_method()` is used to call the next method for the arguments.
This works by looking up the call stack and retrieving R7 methods which have already been called, then doing a method search with those methods excluded.
This ensures you cannot call the same method twice.

```{r}
method(bar, list("text", "number")) <- function(x, y, ...) {
  res <- next_method()(x, y)
  paste0("2 ", res)
}

bar(text("hi"), number(42))
```

## Non-standard evaluation

`method_call()` retains promises for dispatch arguments in basically the same way as `UseMethod()`, so non-standard evaluation works basically the same as S3.

```{r}
subset2 <- new_generic("subset2", signature = "x")

method(subset2, "data.frame") <- function(x, subset = NULL, select = NULL, drop = FALSE, ...) {
  e <- substitute(subset)
  r <- eval(e, x, parent.frame())
  r <- r & !is.na(r)
  nl <- as.list(seq_along(x))
  names(nl) <- names(x)
  vars <- eval(substitute(select), nl, parent.frame())
  x[r, vars, drop = drop]
}

subset2(mtcars, hp > 200, c(wt, qsec))
```

### External generics

If you want to define methods for R7 generics defined in another package you can use `new_extrenal_generic` to declare the external generic, then add `R7::method_register()` to the `.onLoad` function in your package.
`method_register()` will automatically setup on-load hooks for 'soft' dependencies in `Suggests` so the method will be added when the dependency is eventually loaded.

```{r eval = FALSE}
.onLoad <- function(libname, pkgname) {
  R7::method_register()
}

foo <- new_external_generic("pkg1", "foo")

method(foo, list("text", "numeric")) <- function(x, y, ...) paste0("foo-", x, ": ", y)
```

## Design workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

