---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

<!-- badges: start -->
[![R-CMD-check](https://github.com/jimhester/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/jimhester/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/jimhester/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/jimhester/OOP-WG?branch=master)
<!-- badges: end -->

## Workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

## Example

```{r}
library(R7)

range <- class_new("range",
  constructor = function(start, end) {
    object_new(start = start, end = end)
  },
  validator = function(x) {
    if (prop(x, "end") < prop(x, "start")) {
      "`end` must be greater than or equal to `start`"
    }
  },
  properties = c(start = "numeric", end = "numeric")
)

x <- range(start = 1, end = 10)

x@start

x@end

object_class(x)
```

## Performance

The dispatch performance should be roughly on par with S3 and S4, though as this is implemented in the package there is some overhead due to `.Call` vs `.Primitive`.

```{r example}
text <- class_new("text", parent = "character", constructor = function(text) object_new(.data = text))
number <- class_new("number", parent = "numeric", constructor = function(x) object_new(.data = x))

x <- text("hi")
y <- number(1)

foo_r7 <- generic_new(name = "foo_r7", signature = alist(x=))
method_register("foo_r7", "text", function(x) paste0(x, "-foo"))

foo_s3 <- function(x) {
  UseMethod("foo_s3")
}

foo_s3.text <- function(x) {
  paste0(x, "-foo")
}

library(methods)
setOldClass(c("number", "numeric", "r7_object"))
setOldClass(c("text", "character", "r7_object"))

setGeneric("foo_s4", function(x) standardGeneric("foo_s4"))
setMethod("foo_s4", c("text"), function(x) paste0(x, "-foo"))

# Measure performance of single dispatch
bench::mark(foo_r7(x), foo_s3(x), foo_s4(x))


bar_r7 <- generic_new("bar_r7", alist(x=, y=))
method_register("bar_r7", list("text", "number"), function(x, y) paste0(x, "-", y, "-bar"))

setGeneric("bar_s4", function(x, y) standardGeneric("bar_s4"))
setMethod("bar_s4", c("text", "number"), function(x, y) paste0(x, "-", y, "-bar"))

# Measure performance of double dispatch
bench::mark(bar_r7(x, y), bar_s4(x, y))
```
