---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R7

<!-- badges: start -->

[![R-CMD-check](https://github.com/RConsortium/OOP-WG/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RConsortium/OOP-WG/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/RConsortium/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/RConsortium/OOP-WG?branch=master)

<!-- badges: end -->

The R7 package is a new OOP system designed to be a successor to S3 and S4.
It has been designed and implemented collaboratively by the RConsortium Object-Oriented Programming Working Group, which includes representatives from R-Core, BioConductor, RStudio/tidyverse, and the wider R community.

## Installation

The long-term goal of this project is to merge R7 in to base R.
For now, you can experiment with the development version from GitHub:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("r-consortium/OOP-WG")
```

## Usage

This section gives a very brief overview of the entirety of R7.
Learn the basics `vignettte("R7")`, the details of method dispatch in `vignette("dispatch")`, and compatibility with S3 and S4 in `vignette("compatibility")`.

```{r}
library(R7)
```

R7 classes have a formal definition, which includes a list of properties and an optional validator.
Use `new_class()` to define a class:

```{r}
range <- new_class("range",
  properties = list(
    start = class_double, 
    end = class_double
  ),
  validator = function(self) {
    if (length(self@start) != 1) {
      "@start must be length 1"
    } else if (length(self@end) != 1) {
      "@end must be length 1"
    } else if (self@end < self@start) {
      "@end must be greater than or equal to @start"
    }
  }
)
```

`new_class()` returns the class object, which is a constructor that you use to create instances of that class:

```{r}
x <- range(start = 1, end = 10)
x
```

The data possessed by an object is called its **properties**.
Use `@` to get and set properties:

```{r}
x@start
x@end <- 20
x
```

Properties are automatically validated against the type declared in `new_class()` (`double` in this case), and with the class **validator**:

```{r, error = TRUE}
x@end <- "x"
x@end <- -1
```

Like S3 and S4, R7 uses **functional OOP** where methods belong to **generic** functions, and method calls look like all other function calls: `generic(object, arg2, arg3)`.
This style is called functional because from the outside it looks like a regular function call, and internally the components are also functions.

Use `new_generic()` to create a new generic: the first argument is the generic name (used in error messages) and the second gives the arguments used for dispatch.
The third, and optional argument, supplies the body of the generic.
This is only needed if your generic has additional arguments that aren't used for method dispatch.

```{r}
inside <- new_generic("inside", "x", function(x, y) {
  # Actually finds and calls the appropriate method
  R7_dispatch()
})

```

R7 generics support multiple dispatch; see `vignette("dispatch")` for details.

Once you have a generic, you can define a method for a specific class with `method<-`:

```{r}
# Add a method for our class
method(inside, range) <- function(x, y) {
  y >= x@start & y <= x@end
}
inside

inside(x, c(0, 5, 10, 15))
```

You can also use `method<-` to register methods for R7 objects on S3 or S4 generics:

```{r}
method(format, range) <- function(x, ...) {
  paste0("[", x@start, ", ", x@end, "]")
}
format(x)

method(mean, range) <- function(x, ...) {
  (x@start + x@end) / 2
}
mean(x)
```
