---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

<!-- badges: start -->
[![R-CMD-check](https://github.com/jimhester/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/jimhester/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/jimhester/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/jimhester/OOP-WG?branch=master)
<!-- badges: end -->

## Workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

## Example

```{r}
library(R7)

range <- class_new("range",
  constructor = function(start, end) {
    object_new(start = start, end = end)
  },
  validator = function(x) {
    if (prop(x, "end") < prop(x, "start")) {
      "`end` must be greater than or equal to `start`"
    }
  },
  properties = c(start = "numeric", end = "numeric")
)

x <- range(start = 1, end = 10)

x@start

x@end

object_class(x)
```

## Performance

The dispatch performance should be roughly on par with S3 and S4, though as this is implemented in the package there is some overhead due to `.Call` vs `.Primitive`.

```{r example}
text <- class_new("text", parent = "character", constructor = function(text) object_new(.data = text))
number <- class_new("number", parent = "numeric", constructor = function(x) object_new(.data = x))

x <- text("hi")
y <- number(1)

foo_r7 <- generic_new(name = "foo_r7", signature = alist(x=))
method(foo_r7, "text") <- function(x) paste0(x, "-foo")

foo_s3 <- function(x) {
  UseMethod("foo_s3")
}

foo_s3.text <- function(x) {
  paste0(x, "-foo")
}

library(methods)
setOldClass(c("number", "numeric", "r7_object"))
setOldClass(c("text", "character", "r7_object"))

setGeneric("foo_s4", function(x) standardGeneric("foo_s4"))
setMethod("foo_s4", c("text"), function(x) paste0(x, "-foo"))

# Measure performance of single dispatch
bench::mark(foo_r7(x), foo_s3(x), foo_s4(x))


bar_r7 <- generic_new("bar_r7", alist(x=, y=))
method(bar_r7, list("text", "number")) <- function(x, y) paste0(x, "-", y, "-bar")

setGeneric("bar_s4", function(x, y) standardGeneric("bar_s4"))
setMethod("bar_s4", c("text", "number"), function(x, y) paste0(x, "-", y, "-bar"))

# Measure performance of double dispatch
bench::mark(bar_r7(x, y), bar_s4(x, y))
```

## TODO

- Objects
  - [x] - A class object attribute, a reference to the class object, and retrieved with `object_class()`.
  - [x] - For S3 compatibility, a class attribute, a character vector of class names.
  - [ ] - Additional attributes storing properties defined by the class, accessible with @/prop().
- Classes
  - [x] - R7 classes are first class objects with the following
    - [x] - `name`, a human-meaningful descriptor for the class.
      - [ ] - It does _not_ identify the class.
    - [x] - `parent`, the class object of the parent class.
    - [x] - A constructor, an user-facing function used to create new objects of this class. It always ends with a call to `object_new()` to initialize the class.
    - [x] - A validator, a function that takes the object and returns NULL if the object is valid, otherwise a character vector of error messages.
    - [x] - properties, a list of property objects
- Initialization
- Shortcuts
- Validation
  - [ ] - valid_eventually
  - [ ] - valid_implicitly
- Unions
  - [ ] - Used in properties to allow a property to be one of a set of classes
  - [ ] - In method dispatch as a convenience for defining a method for multiple classes
- Properties
  - [x] - Accessed using `prop()` / `prop<-`
  - [x] - Accessed using `@` / `@<-`
  - [x] - A name, used to label output
  - [ ] - A optional class or union
  - [ ] - An optional accessor function
  - [ ] - Properties are created with `prop_new()`
- Generics
  - [x] - It knows its name and the names of the arguments in its signature
  - [x] - Calling `generic_new()` defines a new generic
  - [ ] - By convention, any argument that takes a generic function, can instead take the name of a generic function supplied as a string
- Methods
  - [ ] - 
