---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

These ideas have been implemented in the R7 package, hosted in this repository.

<!-- badges: start -->
[![R-CMD-check](https://github.com/RConsortium/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/RConsortium/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/RConsortium/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/RConsortium/OOP-WG?branch=master)
<!-- badges: end -->


## Classes and objects

```{r error = TRUE}
library(R7)

range <- new_class("range",
  constructor = function(start, end) {
    new_object(start = start, end = end)
  },
  validator = function(x) {
    if (x@end < x@start) {
      "`end` must be greater than or equal to `start`"
    }
  },
  properties = list(
    start = "numeric",
    end = "numeric",
    new_property(
      name = "length",
      class = "numeric",
      getter = function(x) x@end - x@start,
      setter = function(x, value) {
        x@end <- x@start + value
        x
      }
    )
  )
)

x <- range(start = 1, end = 10)

x@start

x@end

x@length

x@length <- 5

x@length

# incorrect properties throws an error
x@middle

# assigning properties verifies the class matches the class of the value
x@end <- "foo"

# assigning properties runs the validator
x@end <- 0

# Print methods for both R7_class objects
object_class(x)

# As well as normal R7_objects
x

# Use `.data` to refer to and retrieve the base data type, properties are
# automatically removed, but non-property attributes (such as names) are retained.

text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

y <- text(c(foo = "bar"))

str(y@.data)
```

## Generics and methods

```{r}
text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

foo <- new_generic(name = "foo", signature = "x")

new_method(foo, list("text"), function(x) paste0("foo-", x))

foo(text("hi"))
```

## Multiple dispatch

```{r}
number <- new_class("number", parent = "numeric", constructor = function(x) new_object(.data = x))

bar <- new_generic(name = "bar", signature = c("x", "y"))

new_method(bar, list("character", "numeric"), function(x, y) paste0("foo-", x, ":", y))

bar(text("hi"), number(42))
```

## Calling the next method

```{r}
new_method(bar, list("text", "number"), function(x, y) {
  res <- next_method()(x, y)
  paste0("2 ", res)
})

bar(text("hi"), number(42))
```

### Load time registration
```{r eval = FALSE}
.onLoad <- function(libname, pkgname) {
  R7::method_register()
}

new_method("pkg1::foo", list("text", "numeric"), function(x, y) paste0("foo-", x, ": ", y))
```

## Performance

The dispatch performance should be roughly on par with S3 and S4, though as this is implemented in the package there is some overhead due to `.Call` vs `.Primitive`.

Dispatch uses a table stored in the `methods` property of the generic. This table is a nested set of hashed environments based on the classes of the methods. e.g.

`method(foo, c("character", "numeric"))` method would be stored at

`foo@methods[["character"]][["numeric"]]`

At each level the search iteratively searches up the class vector for the object.

```{r performance, cache = FALSE}
text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
number <- new_class("number", parent = "numeric", constructor = function(x) new_object(.data = x))

x <- text("hi")
y <- number(1)

foo_R7 <- new_generic(name = "foo_R7", signature = "x")
method(foo_R7, "text") <- function(x) paste0(x, "-foo")

foo_s3 <- function(x) {
  UseMethod("foo_s3")
}

foo_s3.text <- function(x) {
  paste0(x, "-foo")
}

library(methods)
setOldClass(c("number", "numeric", "R7_object"))
setOldClass(c("text", "character", "R7_object"))

setGeneric("foo_s4", function(x) standardGeneric("foo_s4"))
setMethod("foo_s4", c("text"), function(x) paste0(x, "-foo"))

# Measure performance of single dispatch
bench::mark(foo_R7(x), foo_s3(x), foo_s4(x))

bar_R7 <- new_generic("bar_R7", c("x", "y"))
method(bar_R7, list("text", "number")) <- function(x, y) paste0(x, "-", y, "-bar")

setGeneric("bar_s4", function(x, y) standardGeneric("bar_s4"))
setMethod("bar_s4", c("text", "number"), function(x, y) paste0(x, "-", y, "-bar"))

# Measure performance of double dispatch
bench::mark(bar_R7(x, y), bar_s4(x, y))
```

A potential optimization is caching based on the class names, but lookup should be fast without this.

The following benchmark generates a class heiarchy of different levels and lengths of class names and compares the time to dispatch on the first class in the hiearchy vs the time to dispatch on the last class.

We find that even in very extreme cases (e.g. 100 deep heirachy 100 of character class names) the overhead is reasonable, and for more reasonable cases (e.g. 10 deep hiearchy of 15 character class names) the overhead is basically negligible.

```{r performance-2, message = FALSE, R.options = list(width = 120), cache = TRUE}
library(R7)

gen_character <- function (n, min = 5, max = 25, values = c(letters, LETTERS, 0:9)) {
  lengths <- sample(min:max, replace = TRUE, size = n)
  values <- sample(values, sum(lengths), replace = TRUE)
  starts <- c(1, cumsum(lengths)[-n] + 1)
  ends <- cumsum(lengths)
  mapply(function(start, end) paste0(values[start:end], collapse=""), starts, ends)
}

bench::press(
  num_classes = c(3, 5, 10, 50, 100),
  class_size = c(15, 100),
  {
    # Construct a class hierarchy with that number of classes
    text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
    parent <- text
    classes <- gen_character(num_classes, min = class_size, max = class_size)
    env <- new.env()
    for (x in classes) {
      assign(x, new_class(x, parent = parent, constructor = function(text) new_object(.data = text)), env)
      parent <- get(x, env)
    }

    # Get the last defined class
    cls <- parent

    # Construct an object of that class
    x <- do.call(cls, list("hi"))

    # Define a generic and a method for the last class (best case scenario)
    foo_R7 <- new_generic(name = "foo_R7", signature = "x")
    method(foo_R7, cls) <- function(x) paste0(x, "-foo")

    # Define a generic and a method for the first class (worst case scenario)
    foo2_R7 <- new_generic(name = "foo2_R7", signature = "x")
    method(foo2_R7, R7_object) <- function(x) paste0(x, "-foo")

    bench::mark(
      best = foo_R7(x),
      worst = foo2_R7(x)
    )
  }
)
```

And the same benchmark using double-dispatch vs single dispatch

```{r performance-3, message = FALSE, R.options = list(width = 120), cache = TRUE}
bench::press(
  num_classes = c(3, 5, 10, 50, 100),
  class_size = c(15, 100),
  {
    # Construct a class hierarchy with that number of classes
    text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
    parent <- text
    classes <- gen_character(num_classes, min = class_size, max = class_size)
    env <- new.env()
    for (x in classes) {
      assign(x, new_class(x, parent = parent, constructor = function(text) new_object(.data = text)), env)
      parent <- get(x, env)
    }

    # Get the last defined class
    cls <- parent

    # Construct an object of that class
    x <- do.call(cls, list("hi"))
    y <- do.call(cls, list("ho"))

    # Define a generic and a method for the last class (best case scenario)
    foo_R7 <- new_generic(name = "foo_R7", signature = c("x", "y"))
    method(foo_R7, list(cls, cls)) <- function(x, y) paste0(x, y, "-foo")

    # Define a generic and a method for the first class (worst case scenario)
    foo2_R7 <- new_generic(name = "foo2_R7", signature = c("x", "y"))
    method(foo2_R7, list(R7_object, R7_object)) <- function(x, y) paste0(x, y, "-foo")

    bench::mark(
      best = foo_R7(x, y),
      worst = foo2_R7(x, y)
    )
  }
)
```

## Questions

- Best way to support `substitute()` calls in methods? We need to evaluate the argument promises to do the dispatch, but we want to pass the un-evaluated promise to the call?
- Should we remove `method()<-`, We can't do the following, really want to keep this syntax if possible.
  - can't use `method("foo")<-`
  - can't use `method(otherpkg::foo)<-`
  - can't use `method("otherpkg::foo")<-`
- What should happen if you call `new_method()` on a S3 generic?
  1. Should we create a new R7 generic out of the S3 generic?
  2. Or just register the R7 object using `registerS3method()`? ++

## Design workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

## TODO

- Objects
  - [x] - A class object attribute, a reference to the class object, and retrieved with `object_class()`.
  - [x] - For S3 compatibility, a class attribute, a character vector of class names.
  - [x] - Additional attributes storing properties defined by the class, accessible with `@/property()`.
- Classes
  - [x] - R7 classes are first class objects with the following
    - [x] - `name`, a human-meaningful descriptor for the class.
    - [x] - `parent`, the class object of the parent class.
    - [x] - A constructor, an user-facing function used to create new objects of this class. It always ends with a call to `new_object()` to initialize the class.
    - [x] - A validator, a function that takes the object and returns NULL if the object is valid, otherwise a character vector of error messages.
    - [x] - properties, a list of property objects
- Initialization
  - [x] - The constructor uses `new_object()` to initialize a new object, this
    - [x] - Inspects the enclosing scope to find the "current" class.
    - [ ] - Creates the prototype, by either by calling the parent constructor or by creating a base type and adding class and `object_class()` attributes to it.
    - [x] - Validates properties then adds to prototype.
    - [x] - Validates the complete object.
- Shortcuts
  - [ ] - any argument that takes a class object can instead take the name of a class object as a string
  - [x] - instead of providing a list of property objects, you can instead provide a named character vector.
- Validation
  - [x] - valid_eventually
  - [x] - valid_implicitly
- Unions
  - [ ] - Used in properties to allow a property to be one of a set of classes
  - [x] - In method dispatch as a convenience for defining a method for multiple classes
- Properties
  - [x] - Accessed using `property()` / `property<-`
  - [x] - Accessed using `@` / `@<-`
  - [x] - A name, used to label output
  - [ ] - A optional class or union
  - [x] - An optional accessor functions, both getter and setters
  - [ ] - Properties are created with `new_prop()`
- Generics
  - [x] - It knows its name and the names of the arguments in its signature
  - [x] - Calling `new_generic()` defines a new generic
  - [ ] - By convention, any argument that takes a generic function, can instead take the name of a generic function supplied as a string
- Methods
  - Registration
    - [x] - Methods are defined by calling method<-(generic, signature, method):
    - [x] - generic is a generic function.
    - [x] - signature is a
      - [x] - single class object
      - [x] - a class union
      - [x] - list of class objects/unions
      - [x] - a character vector.
    - [ ] - method is a compatible function
    - [x] - `new_method` is designed to work at run-time
      - [ ] - `new_method` should optionally take a package version, so the method is only registered if the package is newer than the version.
    - [ ] - Can define methods where one of the arguments is missing
    - [ ] - Can define methods where one of the arguments has any type
  - Dispatch
    - [x] - Dispatch is nested, meaning that if there are multiple arguments in the generic signature, it will dispatch on the first argument, then the second.
    - [x] - A `plot()` generic dispatching on `x`, e.g. ` plot <- function(x) { method(plot, object_class(x))(x) }`
    - [x] - A `publish()` that publishes an object `x` to a destination `y`, dispatching on both arguments, e.g. `publish <- function(x, y, ...) { method(publish, list(object_class(x), object_class(y)))(x, y, ...) }`
    - [x] - `...` is not used for dispatch
    - [x] - R7 generics can dispatch with base type objects
    - [x] - R7 generics can dispatch with S3 objects
    - [x] - R7 generics can dispatch with S4 objects
    - [x] - `next_method()` can dispatch on multiple arguments, avoiding methods that have already been called.
    - [x] - Generics forward promises to methods, so methods can use non-standard evaluation.
- Compatibility
  - S3
    - [x] - Since the class attribute has the same semantics as S3, S3 dispatch should be fully compatible.
    - [x] - The new generics should also be able to handle legacy S3 objects.
    - [x] - `method()` falls back to single argument S3 dispatch if the R7 dispatch fails.
    - [ ] - `method()` uses S3 group generics as well
  - S4
    - [x] - Since the new generics will fallback to S3 dispatch, they should support S4 objects just as S3 generics support them now.
- Documentation
  - [ ] - Generate index pages that list the methods for a generic or the methods with a particular class in their signature
