---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Object-oriented Programming Working Group

* [Initial proposal](proposal/proposal.org)
* [Requirements brainstorming](spec/requirements.md)
* [Minutes](minutes/)
* [Code](R/) (this repository is an R package)

These ideas have been implemented in the R7 package, hosted in this repository.

<!-- badges: start -->
[![R-CMD-check](https://github.com/RConsortium/OOP-WG/workflows/R-CMD-check/badge.svg)](https://github.com/RConsortium/OOP-WG/actions)
[![Codecov test coverage](https://codecov.io/gh/RConsortium/OOP-WG/branch/master/graph/badge.svg)](https://codecov.io/gh/RConsortium/OOP-WG?branch=master)
<!-- badges: end -->


## Classes and objects

```{r error = TRUE}
library(R7)

range <- new_class("range",
  constructor = function(start, end) {
    new_object(start = start, end = end)
  },
  validator = function(x) {
    if (x@end < x@start) {
      "<range>@end must be greater than or equal to <range>@start"
    }
  },
  properties = list(
    start = "numeric",
    end = "numeric",
    new_property(
      name = "length",
      class = "numeric",
      getter = function(x) x@end - x@start,
      setter = function(x, value) {
        x@end <- x@start + value
        x
      }
    )
  )
)

x <- range(start = 1, end = 10)

x@start

x@end

x@length

x@length <- 5

x@length

# incorrect properties throws an error
x@middle

# assigning properties verifies the class matches the class of the value
x@end <- "foo"

# assigning properties runs the validator
x@end <- 0

# Print methods for both R7_class objects
object_class(x)

# As well as normal R7_objects
x

# Use `.data` to refer to and retrieve the base data type, properties are
# automatically removed, but non-property attributes (such as names) are retained.

text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

y <- text(c(foo = "bar"))

names(y@.data)
```

## Generics and methods

```{r}
text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))

foo <- new_generic(name = "foo", signature = "x")

method(foo, "text") <- function(x, ...) paste0("foo-", x)

foo(text("hi"))
```

## Multiple dispatch

Multiple dispatch uses a table stored in the `methods` property of the generic.
This table is a nested set of hashed environments based on the classes of the methods. e.g.

For `method(foo, c("character", "numeric"))` the method would be stored at `foo@methods[["character"]][["numeric"]]`.

At each level the search iteratively searches along objects class vector.

```{r}
number <- new_class("number", parent = "numeric", constructor = function(x) new_object(.data = x))

bar <- new_generic(name = "bar", signature = c("x", "y"))

method(bar, list("character", "numeric")) <- function(x, y, ...) paste0("foo-", x, ":", y)

bar(text("hi"), number(42))
```

## Calling the next method

`next_method()` is used to call the next method for the arguments.
This works by looking up the call stack and retrieving R7 methods which have already been called, then doing a method search with those methods excluded.
This ensures you cannot call the same method twice.

```{r}
method(bar, list("text", "number")) <- function(x, y, ...) {
  res <- next_method()(x, y)
  paste0("2 ", res)
}

bar(text("hi"), number(42))
```

## Non-standard evaluation

`method_call()` retains promises for dispatch arguments in basically the same way as `UseMethod()`, so non-standard evaluation works basically the same as S3.

```{r}
subset2 <- new_generic(name = "subset", signature = "x")

method(subset2, "data.frame") <- function(x, subset = NULL, select = NULL, drop = FALSE, ...) {
  e <- substitute(subset)
  r <- eval(e, x, parent.frame())
  r <- r & !is.na(r)
  nl <- as.list(seq_along(x))
  names(nl) <- names(x)
  vars <- eval(substitute(select), nl, parent.frame())
  x[r, vars, drop = drop]
}

subset2(mtcars, hp > 200, c(wt, qsec))
```

### External generics

If you want to define methods for R7 generics defined in another package you can use `new_extrenal_generic` to declare the external generic, then add `R7::method_register()` to the `.onLoad` function in your package.
`method_register()` will automatically setup on-load hooks for 'soft' dependencies in `Suggests` so the method will be added when the dependency is eventually loaded.

```{r eval = FALSE}
.onLoad <- function(libname, pkgname) {
  R7::method_register()
}

foo <- new_external_generic("pkg1", "foo")

method(foo, list("text", "numeric")) <- function(x, y, ...) paste0("foo-", x, ": ", y)
```

## Design workflow

* File an issue to discuss the topic and build consensus.
* Once consensus has been reached, the issue author should create a pull 
  request that summarises the discussion in the appropriate `.md` file,
  and request review from all folks who participated the issue discussion.
* Once all participants have accepted the PR, the original author merges.

## TODO

- Objects
  - [x] - A class object attribute, a reference to the class object, and retrieved with `object_class()`.
  - [x] - For S3 compatibility, a class attribute, a character vector of class names.
  - [x] - Additional attributes storing properties defined by the class, accessible with `@/property()`.
- Classes
  - [x] - R7 classes are first class objects with the following
    - [x] - `name`, a human-meaningful descriptor for the class.
    - [x] - `parent`, the class object of the parent class.
    - [x] - A constructor, an user-facing function used to create new objects of this class. It always ends with a call to `new_object()` to initialize the class.
    - [x] - A validator, a function that takes the object and returns NULL if the object is valid, otherwise a character vector of error messages.
    - [x] - properties, a list of property objects
- Initialization
  - [x] - The constructor uses `new_object()` to initialize a new object, this
    - [x] - Inspects the enclosing scope to find the "current" class.
    - [ ] - Creates the prototype, by either by calling the parent constructor or by creating a base type and adding class and `object_class()` attributes to it.
    - [x] - Validates properties then adds to prototype.
    - [x] - Validates the complete object.
- Shortcuts
  - [ ] - any argument that takes a class object can instead take the name of a class object as a string
  - [x] - instead of providing a list of property objects, you can instead provide a named character vector.
- Validation
  - [x] - valid_eventually
  - [x] - valid_implicitly
- Unions
  - [x] - Used in properties to allow a property to be one of a set of classes
  - [x] - In method dispatch as a convenience for defining a method for multiple classes
- Properties
  - [x] - Accessed using `property()` / `property<-`
  - [x] - Accessed using `@` / `@<-`
  - [x] - A name, used to label output
  - [x] - A optional class or union
  - [x] - An optional accessor functions, both getter and setters
  - [x] - Properties are created with `new_property()`
- Generics
  - [x] - It knows its name and the names of the arguments in its signature
  - [x] - Calling `new_generic()` defines a new generic
  - [ ] - By convention, any argument that takes a generic function, can instead take the name of a generic function supplied as a string
- Methods
  - Registration
    - [x] - Methods are defined by calling method<-(generic, signature, method):
    - [x] - generic is a generic function.
    - [x] - signature is a
      - [x] - single class object
      - [x] - a class union
      - [x] - list of class objects/unions
      - [x] - a character vector.
    - [x] - method is a compatible function
    - [x] - `new_method` is designed to work at run-time
      - [x] - `new_method` should optionally take a package version, so the method is only registered if the package is newer than the version.
    - [ ] - Can define methods where one of the arguments is missing
    - [ ] - Can define methods where one of the arguments has any type
  - Dispatch
    - [x] - Dispatch is nested, meaning that if there are multiple arguments in the generic signature, it will dispatch on the first argument, then the second.
    - [x] - A `plot()` generic dispatching on `x`, e.g. ` plot <- function(x) { method(plot, object_class(x))(x) }`
    - [x] - A `publish()` that publishes an object `x` to a destination `y`, dispatching on both arguments, e.g. `publish <- function(x, y, ...) { method(publish, list(object_class(x), object_class(y)))(x, y, ...) }`
    - [x] - `...` is not used for dispatch
    - [x] - R7 generics can dispatch with base type objects
    - [x] - R7 generics can dispatch with S3 objects
    - [x] - R7 generics can dispatch with S4 objects
    - [x] - `next_method()` can dispatch on multiple arguments, avoiding methods that have already been called.
    - [x] - Generics forward promises to methods, so methods can use non-standard evaluation.
- Compatibility
  - S3
    - [x] - Since the class attribute has the same semantics as S3, S3 dispatch should be fully compatible.
    - [x] - The new generics should also be able to handle legacy S3 objects.
    - [x] - `method()` falls back to single argument S3 dispatch if the R7 dispatch fails.
    - [ ] - `method()` uses S3 group generics as well
  - S4
    - [x] - Since the new generics will fallback to S3 dispatch, they should support S4 objects just as S3 generics support them now.
- Documentation
  - [ ] - Generate index pages that list the methods for a generic or the methods with a particular class in their signature
