---
title: "Case Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(R7)
```

## Person class

This case study demonstrates creating basic classes with accessors functions, inheritance and composition.

```{r}
Person <- new_class("Person",
  properties = list(
    name = "character",
    birthdate = new_property(
      class = new_S3_class("Date"),
      setter = function(self, value) {
        self@birthdate <- as.Date(value)
        self
      },
      default = as.Date(NA)
    ),
    age = new_property(
      class = "numeric",
      getter = function(self) {
        # This is not correct because of leap years,
        # see https://stackoverflow.com/questions/3611314/calculate-ages-in-r/25450756#25450756 for a correct implementation.
        as.double(Sys.Date() - self@birthdate) %/% 365
      }
    )
  ),
  validator = function(self) {
    if (length(self@name) != length(self@birthdate)) {
      "@name and @birthdate must be the same length"
    }
  }
)

jane <- Person(name = "Jane Doe")

jane@name

jane@age
```

We can modify objects after creation by assigning to a property.
Unlike in S4 the validity is automatically checked on assignment, not just on object creation.

```{r, error = TRUE}
jane@birthdate <- "1982-03-04"

jane@age

jane@birthdate <- as.Date(c("2001-01-01", "2002-02-02"))
```

Classes can inherit from a parent class with `parent` and contain properties that are of complex classes.

```{r}
Employee <- new_class("Employee",
  parent = Person,
  properties = list(boss = Person)
)

x <- Employee(name = "John Smith", birthdate = "1992-10-11", boss = jane)

x@name

x@age

x@boss
```

## Annotated Data Frame

This case study implements a very basic imitation of a SummarizedExperiment class used widely in the Bioconductor community.

This `annotated_df` class is a normal R `data.frame` with two additional properties
- `rows` which holds a `data.frame` containing data about the _rows_ of the data.
- `columns` which holds a `data.frame` containing data about the _columns_ of the data.

The validator function automatically ensures consistency between the size of the data and the size of the corresponding `rows` and `columns`.
In particular the number of rows of the `rows` property matches the number of rows of the data, and the number of rows of the `columns` property matches the number of columns of the data.

Here is the definition of the class.
```{r}
annotated_df <- new_class("annotated_df",
  properties = list(
    data = new_S3_class("tbl_df"),
    rows = new_S3_class("tbl_df"),
    columns = new_S3_class("tbl_df")
  ),
  validator = function(self) {
    if (nrow(self@data) != nrow(self@rows)) {
      "`nrow(<annotated_df>@rows)` must be equal to `nrow(<annotated_df>)`"
    } else if (ncol(self@data) != nrow(self@columns)) {
      "`nrow(<annotated_df>@columns)` must be equal to `ncol(<annotated_df>)`"
    }
  }
)
```

The print method tells us this is a R7 class and gives us the types for the properties in the class.

```{r}
annotated_df
```

To instantiate an instance of `annotated_df` we first need to get some data to put in it. As an example we will use the mtcars dataset. This requires some manipulation to get the `rows` and `columns` data in the right formats.

```{r}
mt <- mtcars[seq(1, 32, 4), ]

splits <- strsplit(rownames(mt), " ")

mt <- tibble::as_tibble(mt)

vcapply <- function(VALUE, FUN, ...) vapply(VALUE, FUN, character(1), ...)

row_info <- tibble::tibble(
  make = vcapply(splits, head, 1),
  model = vcapply(splits, function(x) paste0(tail(x, -1), collapse = " "))
)

col_desc <- c(
  "mpg" = "Miles/(US) gallon",
  "cyl" = "Number of cylinders",
  "disp" = "Displacement (cu.in.)",
  "hp" = "Gross horsepower",
  "drat" = "Rear axle ratio",
  "wt" = "Weight (1000 lbs)",
  "qsec" = "1/4 mile time",
  "vs" = "Engine (0 = V-shaped, 1 = straight)",
  "am" = "Transmission (0 = automatic, 1 = manual)",
  "gear" = "Number of forward gears",
  "carb" = "Number of carburetors"
)

col_info <- tibble::tibble(name = names(col_desc), description = col_desc)

```

We can then create an instance using the constructor function.

```{r}
obj <- annotated_df(mt, row_info, col_info)
obj@data

obj@rows

obj@columns
```

We next define a `R7_subset` generic function, and then a R7 method for that function, taking integer arguments for the rows and columns.
The subsets the individual components appropriately and then returns a new `annotated_df` object with them.

```{r}
subset_R7 <- new_generic("R7_subset", c("object", "row", "column"))

method(subset_R7, list(annotated_df, "integer", "integer")) <- function(object, row, column, ...) {
  data <- object@data[row, column]
  rows <- object@rows[row, ]
  columns <-  object@columns[column, ]
  annotated_df(data, rows, columns)
}
```

When we subset using this function we can see the consistency is retained for our data components.

```{r}
obj2 <- subset_R7(obj, 1:3, 1:5)

obj2@data

obj2@rows

obj2@columns
```

The validation function is automatically run if you try and change a property, which ensures class integrity.
```{r, error = TRUE}
obj2@rows <- obj2@rows[1:2, ]
```
