---
title: "case_studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{case_studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(R7)
```

## Case Study - Annotated Data Frame

This case study implements a hazy imitation of a SummarizedExperiment class used widely in the Bioconductor community.

This `annotated_df` class is a normal R `data.frame` with two additional properties
- `rows` which holds a `data.frame` containing data about the _rows_ of the data.
- `columns` which holds a `data.frame` containing data about the _columns_ of the data.

The validator function automatically ensures consistency between the size of the data and the size of the corresponding `rows` and `columns`.
In particular the number of rows of the `rows` property matches the number of rows of the data, and the number of rows of the `columns` property matches the number of columns of the data.

Here is the definition of the class.
```{r}
annotated_df <- new_class("annotated_df",
  parent = class_get("data.frame"),
  properties = list(
    rows = "data.frame",
    columns = "data.frame"
    ),
  validator = function(object) {
    if (nrow(object) != nrow(object@rows)) {
      "`nrow(<annotated_df>@rows)` must be equal to nrow(<annotated_df>)"
    } else if (ncol(object) != nrow(object@columns)) {
      "`nrow(<annotated_df>@columns)` must be equal to ncol(<annotated_df>)"
    }
  },
  constructor = function(data, rows, columns) {
    new_object(data, rows = rows, columns = columns)
  }
)
```

The print method tells us this is a R7 class and gives us the types for the properties in the class.

```{r}
annotated_df
```

To instantiate an instance of `annotated_df` we first need to get some data to put in it. As an example we will use the mtcars dataset. This requires some manipulation to get the `rows` and `columns` data in the right formats.

```{r}
mt <- mtcars[seq(1, 32, 4), ]

splits <- strsplit(rownames(mt), " ")

rownames(mt) <- NULL

vcapply <- function(VALUE, FUN, ...) vapply(VALUE, FUN, character(1), ...)

row_info <- data.frame(
  make = vcapply(splits, head, 1),
  model = vcapply(splits, function(x) paste0(tail(x, -1), collapse = " "))
)

col_desc <- c(
  "mpg" = "Miles/(US) gallon",
  "cyl" = "Number of cylinders",
  "disp" = "Displacement (cu.in.)",
  "hp" = "Gross horsepower",
  "drat" = "Rear axle ratio",
  "wt" = "Weight (1000 lbs)",
  "qsec" = "1/4 mile time",
  "vs" = "Engine (0 = V-shaped, 1 = straight)",
  "am" = "Transmission (0 = automatic, 1 = manual)",
  "gear" = "Number of forward gears",
  "carb" = "Number of carburetors"
)

col_info <- data.frame(name = names(col_desc), description = col_desc, row.names = NULL)

```

We can then create an instance using the constructor function.

```{r}
obj <- annotated_df(mt, row_info, col_info)
obj

obj@rows

obj@columns
```

We next define a `R7_subset` generic function, and then a R7 method for that function, taking integer arguments for the rows and columns.
The subsets the individual components appropriately and then returns a new `annotated_df` object with them.

```{r}
subset_R7 <- new_generic("R7_subset", c("object", "row", "column"))

new_method(subset_R7, list(annotated_df, "integer", "integer"), function(object, row, column, ...) {
  data <- object@.data[row, column]
  rows <- object@rows[row, ]
  columns <-  object@columns[column, ]
  annotated_df(data, rows, columns)
})
```

When we subset using this function we can see the consistency is retained for our data components.

```{r}
obj2 <- subset_R7(obj, 1:3, 1:5)

obj2

obj2@rows

obj2@columns
```

The validation function is automatically run if you try and change a property, which ensures class integrity.
```{r, error = TRUE}
obj2@rows <- obj2@rows[1:2, ]
```
