---
title: "Compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

R7 is designed to be compatible with S3 and S4.
This vignette discusses the details.

```{r setup}
library(R7)
```

## S3

R7 objects *are* S3 objects, because R7 is implemented on top of S3.
There are two main differences between an R7 object and an S3 object:

-   As well as the `class` attribute possessed by S3 objects, R7 objects have an additional `R7_class` attribute that contains the object that defines the class.

-   R7 objects have properties; S3 objects have attributes.
    Properties are implemented on top of attributes, so you can access them directly with `attr` and friends.
    When working inside of R7, you should never use attributes directly, but it does mean that existing code will continue to work.

All up, this means most usage of R7 with S3 will just work.

-   R7 can register methods for:

    -   R7 class and S3 generic
    -   S3 class and R7 generic

-   R7 classes can extend S3 classes

-   S3 classes can extend R7 classes

### Methods

`method()` is designed to be the single tool for method registration that you need when working with R7 classes.
You can also register a method for an R7 class and S3 generic without using R7, because all R7 objects have S3 classes, and S3 dispatch will operate on them normally.

```{r}
foo <- new_class("foo")
class(foo())

mean.foo <- function(x, ...) {
  "mean of foo"
}

mean(foo())
```

### Classes

It's possible to extend an R7 class with S3.
This is primarily useful because it many cases it allows you to change a class hierarchy from the inside out: you can provide a formal definition of an S3 class using R7, and its subclasses don't need to change.

### List classes

Many simple S3 classes are implemented as lists, e.g. rle.

```{r}
rle <- function(x) {
  if (!is.vector(x) && !is.list(x)) {
    stop("'x' must be a vector of an atomic type")
  }
  n <- length(x)
  if (n == 0L) {
    new_rle(integer(), x)
  } else {
    y <- x[-1L] != x[-n]
    i <- c(which(y | is.na(y)), n)
    new_rle(diff(c(0L, i)), x[i])
  }
}
new_rle <- function(lengths, values) {
  structure(
    list(
      lengths = lengths, 
      values = values
    ),
    class = "rle"
  )
}
```

There are two ways to convert this to R7.
You could keep the structure exactly the same, using a `list` as the underlying data structure and using a constructor to enforce the structure:

```{r}
new_rle <- new_class("rle", "list", 
  constructor = function(lengths, values) {
    new_object(list(lengths = lengths, values = values))
  }
)
rle(1:10)
```

Alternatively you could convert it to the most natural representation using R7:

```{r}
new_rle <- new_class("rle", properties = list(
  lengths = "integer",
  values = "atomic"
))
```

To allow existing methods to work you'll need to override `$` to access properties instead of list elements:

```{r}
method(`$`, new_rle) <- prop
rle(1:10)
```

The chief disadvantage of this approach is any subclasses will need to be converted to R7 as well.

## S4

R7 properties are equivalent to S4 slots.
The chief difference is that they can be dynamic.

-   R7 classes can not extend S4 classes
-   S4 classes can extend S3 classes
-   R7 can register methods for:
    -   R7 class and S4 generic
    -   S4 class and R7 generic

### Unions

S4 unions are automatically converted to R7 unions.
There's an important difference in the way that class unions are handled in S4 and R7.
In S4, they're handled at method dispatch time, so when you create `setUnion("u1", c("class1", "class2"))`, `class1` and `class2` now extend `u1`.
In R7, unions are handled at method registration time so that registering a method for a union is just short-hand for registering a method for each of the classes.

```{r}
class1 <- new_class("class1")
class2 <- new_class("class2")
union1 <- new_union(class1, class2)

foo <- new_generic("foo", "x")
method(foo, union1) <- function(x) ""
foo
```

R7 unions allow you to restrict the type of a property in the same way that S4 unions allow you to restrict the type of a slot.
