---
title: "Compatibility with S3 and S4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compatibility with S3 and S4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

S7 is designed to be compatible with S3 and S4.
This vignette discusses the details.

```{r setup}
library(S7)
```

## S3

S7 objects *are* S3 objects, because S7 is implemented on top of S3.
There are two main differences between an S7 object and an S3 object:

-   As well as the `class` attribute possessed by S3 objects, S7 objects have an additional `S7_class` attribute that contains the object that defines the class.

-   S7 objects have properties; S3 objects have attributes.
    Properties are implemented on top of attributes, so you can access them directly with `attr` and friends.
    When working inside of S7, you should never use attributes directly, but it does mean that existing code will continue to work.

All up, this means most usage of S7 with S3 will just work.

-   S7 can register methods for S3 generics and S3 classes
-   S7 classes can extend S3 classes
-   S3 classes can extend S7 classes

### Methods

`method()` is designed to be the single tool for method registration that you need when working with S7 classes.
You can also register a method for an S7 class and S3 generic without using S7, because all S7 objects have S3 classes, and S3 dispatch will operate on them normally.

```{r}
Foo := new_class(package = "Bar")
class(Foo())

`mean.Bar::Foo` <- function(x, ...) {
  "mean of foo"
}

mean(Foo())
```

Note that when defining classes in a package, the S3 class name will be `{package}::{class}`, and that means you'll need to use `` ` `` to define S3 methods.

### Generics

If you convert an S3 generic to an S7 generic and want existing S3 methods (typically in other packages) to continue to work, add a `class_any` method that calls `UseMethod()`:

```{r}
foo := new_generic("x")

# Default method dispatches to S3
method(foo, class_any) <- function(x, ...) {
  UseMethod("foo")
}
```

And then if you need a genuine default method, use S3:

```{r}
# Define the "true" default method
foo.default <- function(x) paste0(x, " + 10")
```

### Classes

It's possible to extend an S7 class with S3.
This is primarily useful because in many cases it allows you to change a class hierarchy from the inside out: you can provide a formal definition of an S3 class using S7, and its subclasses don't need to change.

### List classes

Many simple S3 classes are implemented as lists, e.g. rle.

```{r}
rle <- function(x) {
  if (!is.vector(x) && !is.list(x)) {
    stop("'x' must be a vector of an atomic type")
  }
  n <- length(x)
  if (n == 0L) {
    Rle(integer(), x)
  } else {
    y <- x[-1L] != x[-n]
    i <- c(which(y | is.na(y)), n)
    Rle(diff(c(0L, i)), x[i])
  }
}
Rle <- function(lengths, values) {
  structure(
    list(
      lengths = lengths,
      values = values
    ),
    class = "Rle"
  )
}
```

There are two ways to convert this to S7.
You could keep the structure exactly the same, using a `list` as the underlying data structure and using a constructor to enforce the structure:

```{r}
Rle := new_class(
  parent = class_list,
  constructor = function(lengths, values) {
    new_object(list(lengths = lengths, values = values))
  }
)
rle(1:10)
```

Alternatively you could convert it to the most natural representation using S7:

```{r}
Rle := new_class(properties = list(
  lengths = class_integer,
  values = class_atomic
))
```

To allow existing code that uses `$` to keep working, like `inverse.rle()`, you'll need to override `$` to access properties instead of list elements:

```{r}
method(`$`, Rle) <- prop
inverse.rle(rle(1:10))
```

The chief disadvantage of this approach is any subclasses will need to be converted to S7 as well.

## S4

S7 properties play a role similar to S4 slots.
The chief difference is that S7 properties can be dynamic, while S4 slots are
stored fields with a declared S4 class.

S7 and S4 can interoperate in both directions:

-   S7 classes can extend S4 classes.
-   S4 classes can extend S7 classes.
-   S7 can register methods for S4 generics and S4 classes.

### S7 classes that extend S4 classes

To extend an S4 class, supply its class definition or class generator as the
`parent` of a new S7 class. The S4 parent slots become S7 properties on the new
class, and `new_class()` automatically registers the S7 class with S4.

```{r}
methods::setClass("S4Point",
  slots = c(x = "numeric", y = "numeric")
)

ColoredPoint <- new_class("ColoredPoint",
  parent = methods::getClass("S4Point"),
  properties = list(color = class_character)
)

pt <- ColoredPoint(x = 1, y = 2, color = "red")
pt@x
pt@color
methods::is(pt, "S4Point")
```

S4 dispatch follows the S4 parent hierarchy, so S4 methods defined for the
parent can be called on the S7 child.

```{r}
methods::setGeneric("radius", function(x) standardGeneric("radius"))
methods::setMethod("radius", "S4Point", function(x) {
  sqrt(x@x^2 + x@y^2)
})

radius(pt)
```

S4 code that updates an existing object should use `initialize()` rather than
constructing a new object from `class(x)`. The S7 `initialize()` method accepts
stored S7 properties and S4 slots, and later arguments override earlier ones.

```{r}
pt <- methods::initialize(pt, x = 3, color = "blue")
pt@x
pt@color
```

There are a few important caveats:

-   An S7 object that directly extends S4 is represented as an S3 old-class
    object, so `isS4(pt)` is `FALSE`. Use `methods::is()` to ask about S4
    inheritance.
-   The class vector is not scalar. S4 code that needs one primary class name
    should use `class(x)[1L]`, or `methods::class1(x)` when available.
-   `methods::slot()` and `methods::slot<-()` can read and write stored
    properties, but they bypass custom S7 getters, setters, and ordinary S7
    validation. Call `methods::validObject()` after direct slot updates when
    the object should satisfy the S7 contract.
-   Properties with custom getters or setters cannot cross the S4 inheritance
    boundary as S4 slots.
-   `NULL` properties are stored using the same sentinel that S4 uses for
    `NULL` slots. `prop()`, `@`, and `methods::slot()` translate this back to
    `NULL`; raw attributes should be treated as an implementation detail.

### S4 classes that extend S7 classes

Use `S4_register_contains()` when an S4 class should extend an S7 class. It
returns an S4 shim class name for `methods::setClass(contains = )`. The shim
contains the S7 old class, exposes stored S7 properties as formal S4 slots, and
carries the `S7_class` slot that S7 uses for dispatch and validation.

```{r}
Sample <- new_class("Sample",
  properties = list(
    name = class_character,
    score = class_numeric
  ),
  validator = function(self) {
    if (length(self@score) != 1) {
      "@score must have length 1"
    } else if (self@score < 0) {
      "@score must be non-negative"
    }
  }
)

Sample_S4 <- S4_register_contains(Sample)
methods::setClass("ReplicatedSample",
  contains = Sample_S4,
  slots = c(replicate = "integer")
)

rs <- methods::new("ReplicatedSample",
  name = "A",
  score = 3,
  replicate = 1L
)

isS4(rs)
rs@name
prop(rs, "score")
methods::validObject(rs)
```

S7 dispatch still sees the S7 class, even though the concrete object is an S4
object.

```{r}
sample_label <- new_generic("sample_label", "x")
method(sample_label, Sample) <- function(x) {
  paste0(x@name, ": ", x@score)
}

sample_label(rs)
```

The class returned by `S4_register_contains()` is a compatibility layer, not a
user-facing class. Do not instantiate it directly; use it only as a superclass
for another S4 class. Since S4 subclasses are real S4 objects, S4 code can
access the S7 properties as slots. This is necessary for compatibility, but it
also means S4 code can bypass S7 property setters until validation is run.

### Unions

S4 unions are automatically converted to S7 unions when they are used as S7
property classes.
There's an important difference in the way that class unions are handled in S4 and S7.
In S4, they're handled at method dispatch time, so when you create `setUnion("u1", c("class1", "class2"))`, `class1` and `class2` now extend `u1`.
In S7, unions are handled at method registration time so that registering a method for a union is just short-hand for registering a method for each of the classes.

```{r}
Class1 := new_class()
Class2 := new_class()
Union1 <- new_union(Class1, Class2)

foo := new_generic("x")
method(foo, Union1) <- function(x) ""
foo
```

S7 unions allow you to restrict the type of a property in the same way that S4 unions allow you to restrict the type of a slot.
S7 unions can also be registered with `S4_register()` for use in S4 slots and
method signatures.
