---
title: "R7 method dispatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R7 method dispatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette gives more details about method dispatch.
It assumes you're already familiar with the process of how to create a generic and define a method.

```{r setup}
library(R7)
```

## Supplying `fun`

-   Extra required arguments
-   Extra optional arguments
-   ...
-   Method/generic agreement

## Multiple dispatch

### Special sentinels

`any_class()`

`missing_class()`

## `super()`

`next_method()` is used to call the next method for the arguments.
This works by looking up the call stack and retrieving R7 methods which have already been called, then doing a method search with those methods excluded.
This ensures you cannot call the same method twice.

```{r, eval = FALSE}
method(bar, list(text, "double")) <- function(x, y, ...) {
  res <- next_method()(x, y)
  paste0("2 ", res)
}

bar(text("hi"), 42)
```

## External generics

If you want to define methods for R7 generics defined in another package you can use `new_extrenal_generic` to declare the external generic, then add `R7::external_methods_register()` to the `.onLoad` function in your package.
`external_methods_register()` will automatically setup on-load hooks for 'soft' dependencies in `Suggests` so the method will be added when the dependency is eventually loaded.

```{r eval = FALSE}
.onLoad <- function(libname, pkgname) {
  R7::external_methods_register()
}

foo <- new_external_generic("pkg1", "foo")
method(foo, "integer") <- function(x) paste0("foo-", x)
```

## Internals

Multiple dispatch uses a table stored in the `methods` property of the generic.
This table is a nested set of hashed environments based on the classes of the methods.
e.g.

For `method(foo, c("character", "numeric"))` the method would be stored at `foo@methods[["character"]][["numeric"]]`.

At each level the search iteratively searches along objects class vector.

```{r, eval = FALSE}
bar <- new_generic("bar", c("x", "y"))
method(bar, list("character", "double")) <- function(x, y) paste0("foo-", x, ":", y)

bar("hi", 42)
```

## 
