---
title: "Evolving S7 classes and generics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evolving S7 classes and generics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--
The failures described in this vignette can be verified with `tools/evolution/`: a small harness that builds real package pairs and observes what breaks. Use it to reverify results when the package changes.
-->

Sooner or later every successful package needs to change its interface: a generic gains an argument, a property turns out to be misnamed, a class needs to be split in two.
When other packages depend on yours, these changes stop being private decisions.
This vignette works through the ways an S7 interface can change, and what each change does to downstream packages.

Throughout, we'll imagine two packages:

* **myPackage** is upstream: it defines S7 classes and generics.
  You're the author of myPackage.
* **depPackage** is downstream: it depends on myPackage, registering methods for myPackage's generics, subclassing myPackage's classes, and creating and manipulating instances of them.
  Someone else is the author of depPackage.

Because it's a lot of hassle to release CRAN packages simultaneously, if myPackage needs to make breaking changes, it must be possible to make a version of depPackage that works with both old and new myPackage.
So for each change we care about three questions:

1. What breaks in depPackage, and *when*: when depPackage is installed, when it's loaded, or when it's used?
2. What should the author of depPackage do so that a single version of depPackage works with both versions of your package?
3. What can you do to make the change less disruptive in the first place?

```{r setup}
library(S7)
```

## Designing for evolution

Most of the advice in this vignette boils down to a few design rules that are cheap on day one and pay for themselves the first time you need to change anything:

* **Give every generic `...`** (the default).
  A generic without `...` can never gain an argument without breaking every downstream method.

* **Keep generics minimal.**
  Every non-dispatch argument in the generic's signature is part of a contract that all methods are checked against, and its default belongs to the generic: dispatch supplies it, so no method can override it.
  When an argument is only meaningful to some methods, leave it out of the generic and let those methods take it via `...`.

* **Treat names as permanent.**
  Generics can be renamed with `deprecated_generic()`.
  For classes, `deprecated_class()` preserves the old class and can recommend a replacement; changing a class's identity still affects saved objects and already-installed subclasses.
  Spend naming effort accordingly, and set `package` once, correctly.

* **Pass properties by name** to constructors, always.

* **Prefer composition for volatile details.**
  Fields you expect to churn can live in a helper class stored as one property, so the enclosing class's contract — and every subclass — is insulated from the churn.

* **Deprecate, then remove.**
  Almost every change below has a two-release shape: one release that supports both forms (deprecated alias, deprecated property, warning validator), and a later one that drops the old form.
  Your reverse dependency checks tell you when it's safe to take the second step.

And when a change genuinely can't be made compatible — dispatch changes, class renames, property narrowing — the best tool is social, not technical: work with your downstream maintainers, send PRs, and sequence the releases.

## How S7 couples packages

To predict what breaks, you need to know when downstream code actually runs:

* **Package code runs at install time.**
  depPackage's calls to `new_class()` and `method<-` run when depPackage is *installed*, and their results are saved into depPackage's namespace.
  So a change to myPackage can break depPackage's installation, which is the most severe kind of failure: `R CMD check` fails immediately, and so does every other package that depends on depPackage.

* **Method registrations are replayed at load time.**
  When depPackage registers a method for a generic in myPackage, the registration is also replayed every time depPackage is loaded (that's what `S7_on_load()` does; see `vignette("packages")`).
  So a method that depPackage recorded against the old myPackage is re-registered against whatever myPackage is installed now.

* **Everything else happens at run time.**
  Calls through `myPackage::Foo()` and external references use the installed myPackage, but objects and class definitions saved in depPackage can retain parts of the old definition.
  Failures here don't break installation; they surface wherever the code runs, most visibly in depPackage's tests, examples, and vignettes.

### The generic–method contract

When a method is registered, S7 compares the method's formals to the generic's:

* The method's first arguments must match the generic's dispatch arguments, in order, with no defaults.
  A mismatch is an **error**.
* If the generic doesn't have `...`, the method's formals must match the generic's exactly.
  Any difference is an **error**.
* If the generic has a non-dispatch argument that the method lacks, that's a **warning**.
* If the method's default for an argument differs from the generic's, that's a **warning**.
* If the method has extra arguments that the generic lacks (and the generic has `...`), that's **silently allowed**.

That last rule is the single most useful tool for coordinating change, because it means a method can be written for a *future* version of its generic.
We'll use it repeatedly below.

These checks only run where someone can act on them, i.e. in **development contexts**:

* Under `devtools::load_all()`.
* Under `R CMD check`, but only when the package being checked owns the method, the generic, or one of the classes in the signature.
* When the method isn't in a package at all — registered in a script or interactively, as in this vignette's examples.

Everywhere else the checks are skipped and the method is registered as-is: plain `R CMD INSTALL`, an ordinary `library()` call, and `R CMD check` of some *other* package that merely depends on depPackage.
This shapes almost everything below.
When myPackage breaks the contract, the mismatch is reported to depPackage's author, and to you when you run reverse dependency checks (those check depPackage itself), but not to depPackage's users, and not to depPackage's own downstream packages.

The price is that the two **error** cases no longer stop an end user from installing or loading depPackage.
The mismatched method is registered anyway, and the failure resurfaces at run time when the generic is called.

One extra affordance while `load_all()` is active: the errors are demoted to warnings, and the offending method isn't registered.
That means you can re-source a package whose generic you just changed and fix its methods one at a time, instead of having to update them all before the package will load again.

### How depPackage should refer to myPackage

One mechanical detail matters more than you might expect: `method<-` is a replacement function, so `method(myPackage::gen, MyClass) <- f` doesn't work — R can't assign to `myPackage::gen`.
depPackage has three options, and they behave differently when myPackage changes:

* **Import the generic** (`@importFrom myPackage gen`, then `method(gen, MyClass) <- f`).
  This is the standard approach for a hard dependency.
  If `gen` disappears from myPackage, depPackage fails to *install*.

* **Use `new_external_generic()`**.
  This registers the method only when myPackage is loaded, and optionally only when myPackage is new enough (via `version`).
  If the generic disappears, you get a warning at load time rather than an installation failure.

* **Use `new_external_class()`** to refer to myPackage's classes, both in method signatures and as the `parent` of a subclass.
  The subclass's constructor forwards `...` to the installed `myPackage::Foo()` instead of baking in its arguments and defaults.
  The subclass still stores property definitions and dispatch metadata from installation, so structural changes can require rebuilding depPackage.

```{r, eval = FALSE}
# In depPackage
Foo := new_external_class(package = "myPackage")
MyFoo := new_class(parent = Foo, properties = list(extra = class_double))

gen := new_external_generic(package = "myPackage", dispatch_args = "x")
method(gen, MyFoo) <- function(x, ...) "Hello!"
```

### Signaling deprecation

S7 supplies three helpers for keeping an old interface available during a transition:

| Helper | With a replacement | Without a replacement |
|--------|--------------------|-----------------------|
| `deprecated_generic()` | Calls warn and dispatch through `new`; registrations target `new` silently. | Replace `new_generic()` in the existing definition, preserving its dispatch arguments and function. Calls warn and still dispatch. |
| `deprecated_class()` | Define the original class and supply `replacement` to recommend another class. Constructor calls warn and still create the original class. | Omit `replacement`; explicit constructor calls warn. |
| `deprecated_property()` | Reads and changed writes warn and use the property named by `new`. | Omit `new` to retain storage and warn on reads and changed writes. |

All three require `when`, the version of your package that introduced the deprecation, and accept the same `method` argument:

* `"base"`, the default, signals a `deprecatedWarning` on each deprecated call or access that triggers the helper.
* `"lifecycle(warn)"` uses `lifecycle::deprecate_warn()`, which throttles repeated warnings.
* `"lifecycle(stop)"` uses `lifecycle::deprecate_stop()` to make that use an error.

If you use a lifecycle method, add lifecycle to your package's dependencies.
The warning policy is separate from registration: even `"lifecycle(stop)"` permits method registration and use of a deprecated class in class definitions.
It stops calls to the deprecated constructor or generic, or access to the deprecated property, rather than preventing downstream packages from referring to the definition.

Warnings in a plain R test script do not by themselves make `R CMD check` fail.
They can still fail tests that require silence or convert warnings to errors, and a warning during namespace loading can produce a check NOTE.
Use the warning phase to find downstream uses before switching to errors or removing an export.

Downstream packages that save a copy of a generic or class in their namespace can keep using the old definition without warning until they are rebuilt.
Calls that look up the export dynamically, such as `myPackage::gen()`, use the updated definition immediately.

## Changing a generic

### Adding an argument

If your generic has `...` (and it should — see [Designing for evolution]), adding a non-dispatch argument is one of the mildest changes you can make.
Existing methods keep working, but every registration produces a warning until they're updated:

```{r}
# myPackage, version 1
gen := new_generic("x")
# depPackage's method, written against version 1
gen_BClass <- function(x, ...) "old"

# myPackage, version 2 adds `verbose`
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) S7_dispatch())

BClass := new_class()
method(gen, BClass) <- gen_BClass
```

Because the check only runs in development contexts, depPackage's users see nothing: installing and loading depPackage are both silent, and calls keep working.
The warning surfaces under `load_all()`, and as a single NOTE in `R CMD check` of depPackage:

```
* checking whether the namespace can be loaded with stated dependencies ... NOTE
Warning: gen(<depPackage::BClass>) doesn't have argument `verbose`
```

So depPackage's author will hear about it — and so will you, when you run reverse dependency checks, since those check depPackage itself.

The fix is pleasingly asymmetric: because methods may have arguments the generic lacks, depPackage can add the argument **before** myPackage releases:

```{r}
# myPackage, version 1 again
gen := new_generic("x")

# depPackage adds `verbose` ahead of myPackage's release: no warning under version 1...
method(gen, BClass) <- function(x, verbose = FALSE, ...) "new"

# ...and none under version 2 either
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) S7_dispatch())
method(gen, BClass) <- function(x, verbose = FALSE, ...) "new"
```

So the transition recipe is:

1. You tell downstream authors about the upcoming argument (or send PRs).
2. Each downstream package adds the argument to its methods, and can release whenever convenient.
3. You release.

If you skip steps 1–2, nothing breaks outright — downstream packages just get a check NOTE until they catch up.

If your generic *doesn't* have `...`, adding an argument is an error for every downstream method, and there is no version of the method that satisfies both the old and the new generic:

```{r, error = TRUE}
gen_strict := new_generic("x", fun = function(x) S7_dispatch())
method(gen_strict, BClass) <- function(x) "old"

# myPackage, version 2
gen_strict := new_generic("x", fun = function(x, verbose = FALSE) S7_dispatch())
method(gen_strict, BClass) <- function(x) "old"
```

Since that error is only raised in development contexts, depPackage still installs and loads for its users; the mismatched method is registered, and the call blows up when it's made:

```
Error in `method(gen_strict, depPackage::BClass)`(x = <object>, verbose = FALSE) :
  unused argument (verbose = FALSE)
```

depPackage's own `R CMD check` fails earlier and says something more useful, at `checking whether the package can be loaded`.
Either way there is no fix on depPackage's side, so a generic without `...` is a frozen interface: only add one deliberately, and treat any change to it like a rename (below).

### Removing an argument

Removing a non-dispatch argument is silent at registration time: methods may have extra arguments, so a method that still lists the removed argument registers without complaint against both versions.
Downstream packages can remove the argument from their methods at leisure.

Callers are the bigger concern: `gen(x, removed_arg = 1)` will still be accepted by the generic (the argument passes through `...`) and then reach methods that may or may not expect it.
So before removing an argument, deprecate it: keep it in the generic, and warn when it's supplied.
This is an argument deprecation, so put the warning in the generic's function rather than wrapping the whole generic with `deprecated_generic()`:

```{r}
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) {
  if (!missing(verbose)) {
    warning("`verbose` is deprecated.")
  }
  S7_dispatch()
})
method(gen, BClass) <- function(x, verbose = FALSE, ...) "result"
gen(BClass())
gen(BClass(), verbose = TRUE)
```

Use `lifecycle::deprecate_warn()` in the same place if your package uses lifecycle.
Keep the method's argument during the transition so it still matches the generic.

### Changing a default

Changing the default value of a non-dispatch argument warns for every method whose default doesn't match — and, more consequentially, changes what those methods actually receive.
The default belongs to the generic, not to the method: dispatch passes on all of the generic's arguments, so when a caller supplies nothing, the method gets the *generic's* default and its own is never reached.

```{r}
gen := new_generic("x", fun = function(x, drop = TRUE, ...) S7_dispatch())
method(gen, BClass) <- function(x, drop = FALSE, ...) drop
gen(BClass())
```

(A method's default is only used when the generic's argument has no default at all.)
So changing a default changes behavior immediately, for every caller — including callers of methods that thought they had pinned the old value.

Unlike an added argument, there's no spelling of the method that agrees with both versions of myPackage: whichever default it names disagrees with one of them, and omitting the default disagrees with both.
So the mitigation here isn't a transition recipe, it's the scoping described in [The generic–method contract]: the warning fires only in development contexts, which costs depPackage one check NOTE and costs depPackage's users nothing.

Our advice: change defaults rarely, and prefer to keep non-dispatch arguments (and especially their defaults) out of the generic entirely, so that the methods that care about them own both the argument and its default.

### Renaming a generic

If you simply rename a generic, depPackage fails to install: its `importFrom(myPackage, gen)` directive (or `myPackage::gen` call) no longer resolves.

The fix is to keep the old name as a deprecated alias for the new generic:

```{r}
# myPackage, version 2
gen2 := new_generic("x")
gen1 := deprecated_generic(new = gen2, when = "2.0.0") # export both
method(gen1, BClass) <- function(x, ...) "result"
gen2(BClass())
gen1(BClass())
```

`deprecated_generic()` returns an object that still counts as the generic for registration purposes — `method(gen1, MyClass) <- f` lands on the shared `gen2`, so depPackage keeps installing and both names dispatch to the same methods — while calls through the old name warn that it's time to move on.
This also works for already-installed depPackage registrations and for registrations made with `new_external_generic()`.
(If you plan to keep the old name indefinitely, a plain alias `gen1 <- gen2` is completely transparent and signals nothing.)

An ordinary wrapper such as `function(x, ...) { .Deprecated("gen2"); gen2(x, ...) }` is not a generic that downstream packages can register methods on.
Use `deprecated_generic()` when the old name is an extension point.

You can also keep the generic's internal identity while changing its exported name.
Use `new_label` when the name in the warning should differ from the generic's internal name:

```{r}
gen1 := new_generic("x")
gen2 <- gen1
gen1 := deprecated_generic(new = gen2, when = "2.0.0", new_label = "gen2()")
method(gen1, BClass) <- function(x, ...) "result"
gen1(BClass()) # warns and recommends gen2()
gen2(BClass())
```

`new_label` changes only the message; both exports use the same generic and methods.

### Retiring a generic

If there is no replacement, change `new_generic()` to `deprecated_generic()` in the generic's definition and add `when`:

```{r}
retired := deprecated_generic("x", when = "2.0.0")
method(retired, BClass) <- function(x, ...) "result"
retired(BClass())
```

Keep the existing `dispatch_args`, any custom `fun`, and the method registrations.
The old name and dispatch signature stay available, so downstream packages can continue registering methods during the warning phase.
Removing the export later still breaks imports; changing to `method = "lifecycle(stop)"` instead preserves registration but makes calls fail.

### Changing dispatch

Changing a generic's dispatch arguments — renaming them, reordering them, or moving between single and multiple dispatch — invalidates every downstream method and method registration, so downstream packages fail to install.
There is no transition trick: a generic's dispatch arguments are its identity.
If you need different dispatch, create a generic with a new name and keep the old generic during the transition.
`deprecated_generic(new = ...)` adopts the replacement's formals and dispatch signature; it does not translate an old registration or adapt arguments to a different interface.
Use `deprecated_generic()` with the old generic's definition while you coordinate new methods, and write any argument adaptation explicitly.

### Moving a generic to another package

Splitting generics out into a lower-dependency package (say myPackageCore) works, as long as myPackage re-exports them *through its `NAMESPACE`*:

```{r, eval = FALSE}
# myPackage, version 2
#' @importFrom myPackageCore gen
#' @export
myPackageCore::gen
```

depPackage's `importFrom(myPackage, gen)` still resolves, and method registration follows the generic object to its new home package automatically, so depPackage works without any changes — even an already-*installed* depPackage, whose deferred registrations resolve through the re-export the next time it's loaded.

There is a trap here: a binding *copy* (`gen <- myPackageCore::gen` in myPackage's code, the idiom that works fine for a within-package rename) is not a re-export.
It runs when myPackage is installed, and serializes a *duplicate* of the generic — with its own private methods table — into myPackage's namespace.
The result is a split brain: depPackage's registrations follow the generic's home package and land on myPackageCore's original, while calls through `myPackage::gen()` dispatch using the duplicate's empty table and fail with `Can't find method`.
A `NAMESPACE` re-export resolves to the one true generic at load time, so use exactly that.

Should the re-export signal deprecation?
Since a re-export costs nothing to maintain, keeping it forever is a fine choice.
Keep the `NAMESPACE` re-export while downstream packages migrate to the new home.

If you want to retire the old home, export a deprecated alias instead:

```{r, eval = FALSE}
# In myPackage, version 2
gen := deprecated_generic(new = myPackageCore::gen, when = "2.0.0")
```

The wrapper resolves the generic in its owning namespace, so calls and method registrations through either package use the same method table.
This works for both already-installed and rebuilt downstream packages.
The target must remain available under its original name in that namespace.

## Changing a class

For everything below, we assume depPackage subclasses via `new_external_class()`, as recommended above.
Where the behavior differs for `parent = myPackage::Foo` (a direct reference to the class object), we say so.

### Adding a property

Adding a property is usually compatible with downstream source: subclasses pick it up when depPackage is rebuilt.
An external parent's constructor resolves at run time, but an already-installed subclass still stores the property definitions it was built with.
Do not assume that forwarding constructor arguments also refreshes that metadata.

There are two ways it can still bite:

* **Name clash.** If depPackage's subclass already has a property with the same name and its type doesn't narrow yours, depPackage fails to install:

  ```
  Error: <MyFoo>@y must narrow <myPackage::Foo>@y.
  ```

  depPackage then has to rename its property — which is itself a property rename with its own deprecation cycle (below).
  There's no general defense, but clashes are less likely if subclass authors avoid very generic property names on other people's classes, and if you announce new properties ahead of time.

* **Positional construction.** `myPackage::Foo(1, 2)` changes meaning if a property is inserted before an existing one.
  Downstream code should always pass properties to constructors by name; upstream, only append properties.

### Removing a property

Removing a property doesn't break installation — nothing about a property is checked when depPackage is installed unless depPackage overrides it — but every use of it breaks at run time: `Foo(count = 2)` fails with `unused argument`, and `x@count` fails with `Can't find property`.
These failures surface in depPackage's tests, examples, and vignettes.

To give downstream packages a transition window, keep the property with `deprecated_property()` and omit `new`:

```{r}
# myPackage, version 1
validate_count <- function(value) {
  if (any(value < 0)) "must be non-negative"
}
Foo := new_class(properties = list(
  count = new_property(class_double, default = 1, validator = validate_count)
))

# myPackage, version 2
Foo := new_class(properties = list(
  deprecated_property(
    "count", when = "2.0.0", class = class_double, default = 1,
    validator = validate_count
  )
))

foo <- Foo(count = 3)
foo@count
foo@count <- 4
print(foo)
```

Preserve the original property's class, default, and validator explicitly.
Invalid values still fail validation during construction and assignment.
The helper does not wrap an existing getter or setter; retiring a computed property or one with a custom setter needs a custom recipe that preserves that behavior.
Without a replacement, construction does not warn, including when `count` is supplied; reads and changed writes do.
The default `print()` and `str()` omit deprecated properties without calling their getters, including when access would signal an error.
`props()` still reads them and therefore still signals deprecation.

An already-installed subclass can keep using its old property definition silently.
Maintainers need to rebuild the subclass package against the updated parent to adopt the deprecation; users then need to install that rebuilt version.
Changing the parent package alone does not guarantee that reads and writes in every subclass will warn.

### Renaming a property

A rename is "add the new property, deprecate the old one":

```{r}
Foo := new_class(properties = list(
  size = new_property(class_double, default = 1),
  deprecated_property("count", new = "size", when = "2.0.0")
))

foo <- Foo(size = 3) # silent
foo <- Foo(count = 3) # warns, then sets size
foo@count
foo@count <- 4
foo@size
```

The helper supplies both a getter and setter, so callers can keep using `Foo(count = ...)`, `foo@count`, and `foo@count <- ...` during the transition.
Put the replacement property first and preserve its old default and validation rules.
Put any `validator` on the replacement property; `deprecated_property()` rejects a validator when `new` is supplied.
The deprecated argument defaults to the replacement argument; callers using only `size` stay silent.
If both names are supplied with different values, the deprecated argument wins and warns.
Avoid supplying both names.

There is a signaling limit: an explicit assignment of the current value can be silent, as can construction with equal `count` and `size`.
The helper uses value equality to avoid warning during default construction, so it cannot distinguish those explicit uses from initialization.
This also bypasses `"lifecycle(stop)"` for those writes; it is not a complete ban on supplying the old name.
Once downstream packages have migrated, remove the deprecated property.

An already-installed subclass needs rebuilding for this transition.
Its saved definition can still describe `count` as a stored property, while the new parent stores only `size`.
In the compatibility lab, constructing that stale subclass fails validation because its `count` storage is missing; the same downstream source works after rebuilding.

### Changing a property's type or validator

Widening a property's type (say, `class_integer` to `class_numeric`) is safe for downstream code: everything that validated before still validates.
If a subclass overrides the property, its override must narrow yours; an override written against the old, narrower type still narrows the widened type, so it keeps installing.

Narrowing is a run-time break: depPackage installs fine, but any code that stored a now-invalid value fails when it runs:

```
Error: <myPackage::Foo> object properties are invalid:
- @size must be <integer>, not <double>
```

The same applies to tightening a validator.
There's no mechanical transition here; treat narrowing like removing the property and adding a stricter one, i.e. give downstream packages a deprecation window in which the validator warns rather than errors.

### Deprecating a class

Change `new_class()` to `deprecated_class()` in the existing class definition and add `when`.
Keep the class's name, package, parent, properties, constructor, and validator unchanged.
If there is a replacement, supply it as `replacement`:

```{r}
# myPackage, version 1
Foo := new_class(properties = list(size = class_double))
saved <- Foo(size = 1)
MyFoo := new_class(parent = Foo)
measure := new_generic("x")
method(measure, Foo) <- function(x) x@size

# myPackage, version 2: export both classes
Bar := new_class(properties = list(size = class_double))
Foo := deprecated_class(
  properties = list(size = class_double),
  replacement = Bar,
  when = "2.0.0"
)
foo <- Foo(size = 2) # warns and still constructs a Foo
measure(foo)
measure(MyFoo(size = 3)) # existing subclasses still work, without warning
measure(saved)
S7_inherits(saved, Foo)
S7_inherits(saved, Bar)
```

`replacement` changes the message only.
`Foo` keeps its identity, and existing methods still apply to `Foo`, its subclasses, and saved instances.
Adding deprecation to an otherwise unchanged class does not require rebuilding packages that define subclasses, whether their parent reference is direct or external.
Changes to the class's definition still need the compatibility work described in the other sections.

Methods for `Foo` do not automatically apply to `Bar`.
When a method can handle both classes, register it on their union:

```{r}
method(measure, Foo | Bar) <- function(x) x@size
measure(Bar(size = 4))
measure(saved)
```

If the replacement needs a different representation, keep separate methods and provide an explicit conversion for users who want it.
Recommending `Bar` does not convert existing objects or change the type returned by `Foo()`.

Class contexts remain available without warnings: method signatures, parent classes, property types, unions, and `new_external_class()` references use the original class.
Generated subclass constructors also retain the original constructor's defaults and validation.
Even `method = "lifecycle(stop)"` permits these uses; it stops explicit calls to the deprecated constructor.

Property defaults have one installation boundary.
A previously installed class defined with `properties = list(item = myPackage::Foo)` can have a generated default that calls `myPackage::Foo()` directly.
After myPackage deprecates `Foo`, using that default warns, or errors in stop mode, until the containing package is rebuilt.
Defaults generated from `new_external_class()` resolve the underlying class silently, including in already-installed packages.
Rebuilding a direct property reference against the deprecated class also generates a silent default.
An explicitly written default such as `quote(myPackage::Foo())` remains an ordinary constructor call and signals deprecation.

### Renaming or moving a class

A class's identity — the string `"myPackage::Foo"` — appears in method dispatch tables, existing instances, and installed subclass definitions.
To introduce a class named `Bar`, keep `Foo` available with its existing definition and deprecate it as above.
Downstream authors can then move their subclasses and methods to `Bar` when ready.

For a class defined with `new_class()`, adding `Bar <- Foo` gives it another exported spelling while preserving its identity and constructor behavior.
`deprecated_class()` defines a class; it does not wrap an existing class as an alias, and it does not accept `new`, `old`, or `new_label`.
The `new_label` argument belongs to `deprecated_generic()`.

A plain alias in the other direction, `Foo <- Bar` after defining a new class named `Bar`, preserves the old export but changes which class it names.
External references can resolve through this alias, so unchanged downstream source can be rebuilt.
An already-installed subclass still has `myPackage::Foo` in its dispatch vector while a method registered through the old export now targets `myPackage::Bar`.
The lab verifies that dispatch fails until the subclass package is rebuilt.
Saved instances retain their old identity and need explicit conversion or reconstruction if they are to become `Bar` objects.

Changing the class's `package` also changes its identity.
To recommend a class in another package while preserving existing users, keep the original definition:

```{r, eval = FALSE}
# In myPackage
Foo := deprecated_class(
  properties = list(size = class_double),
  replacement = myPackageCore::Foo,
  when = "2.0.0"
)
```

The warning recommends `myPackageCore::Foo()`, but construction and methods still use `myPackage::Foo`.
Actually moving the class and exporting `Foo <- myPackageCore::Foo` has the same rebuilding and saved-object limits as a rename that changes identity.
Removing the old export entirely breaks downstream imports and class references.

### Retiring a class

If there is no replacement, omit `replacement` and keep the existing definition:

```{r}
Retired := deprecated_class(
  properties = list(size = class_double),
  when = "2.0.0"
)
Retired(size = 3)
```

Downstream subclasses, method signatures, property types, and unions can continue using the class; explicit constructor calls warn.
If you intend to remove the class entirely, check these downstream definitions as well as direct constructor calls.

### Making a class abstract

Making an existing class abstract breaks every direct `myPackage::Foo()` call at run time, which is presumably the point of the change — give downstream packages a deprecation window by replacing `new_class()` with `deprecated_class()` in its definition, and only later make the class abstract.

Subclasses are unaffected, including already-installed ones: even though a subclass with an external parent builds its parent part by calling `myPackage::Foo()` at run time, an abstract class's constructor is allowed to run when it's constructing the parent part of a concrete subclass.
