---
title: "Evolving S7 classes and generics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evolving S7 classes and generics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--
The failures described in this vignette can be verified with `tools/evolution/`: a small harness that builds real package pairs and observes what breaks. Use it to reverify results when the package changes.
-->

Sooner or later every successful package needs to change its interface: a generic gains an argument, a property turns out to be misnamed, a class needs to be split in two.
When other packages depend on yours, these changes stop being private decisions, and you need to think through how your changes affect your dependencies.
This vignette works through the ways an S7 interface can change, and what each change does to downstream packages. 
We assume that you're already familiar with `vignette("packages.Rmd")`, although we'll repeat the most important content.

Throughout, we'll imagine two packages:

* **myPackage** is upstream: it defines S7 classes and generics.
  You're the author of myPackage.
* **depPackage** is downstream: it depends on myPackage, registering methods for myPackage's generics, subclassing myPackage's classes, and creating and manipulating instances of them.
  Someone else is the author of depPackage.

Because it's a lot of hassle to release CRAN packages simultaneously, if myPackage needs to make breaking changes, it must be possible to make a version of depPackage that works with both old and new myPackage.
So for each change we care about three questions:

1. What breaks in depPackage, and *when*: when depPackage is installed, when it's loaded, or when it's used?
2. What should the author of depPackage do so that a single version of depPackage works with both versions of your package?
3. What can you do to minimise the disruption of the change?

```{r setup}
library(S7)
```

## Designing for evolution

You can make life easier for yourself and your dependencies by following a few design rules that are cheap during development and pay for themselves as soon as you need to change anything:

* **Give every generic `...`** (the default).
  A generic without `...` can never gain an argument without breaking every downstream method.

* **Keep generics as small as possible.**
  Every argument in the generic's signature (even those that don't take part in dispatch) is part of a contract that all methods are checked against.
  When an argument is only meaningful to some methods, leave it out of the generic and let those methods take it via `...`.

* **Treat names as permanent.**
  While there are ways to "rename" classes and generics, it's best to think of them as creating a new class/generic and deprecating the old. 

* Always **pass constructor properties by name**, not position.

* **Prefer composition for volatile details.**
  Put fields that you expect to churn in a helper class stored as one property, insulating subclasses from the churn.

## How S7 couples packages

To predict what breaks, you need to know when downstream code actually runs:

* **Package code runs at install time.**
  depPackage's calls to `new_class()` and `method<-` run when depPackage is *installed*, and their results are saved into depPackage's namespace.
  So a change to myPackage can break depPackage's installation, which is the most severe kind of failure: `R CMD check` fails immediately, and so does every other package that depends on depPackage.

* **Method registrations are replayed at load time.**
  When depPackage registers a method for a generic in myPackage, the registration is also replayed every time depPackage is loaded (that's what `S7_on_load()` does; see `vignette("packages")`).
  So a method that depPackage recorded against the old myPackage is re-registered against whatever myPackage is installed now.

* **Everything else happens at run time.**
  Calls through `myPackage::Foo()` and external references use the installed myPackage, but objects and class definitions saved in depPackage can retain parts of the old definition.
  Failures here don't break installation; they surface wherever the code runs, most visibly in depPackage's tests, examples, and vignettes.

It's most important to avoid failures that happen at install time, because these make it impossible to create a version of depPackage that works with both old and new myPackage, thus requiring simultaneous CRAN releasese.
This is a significant cost, and will require carefully coordination with downstream maintainers.

### Developing vs using

In fact, it's so important to avoid problems at install time, that S7 surfaces errors differently to developers vs end-users.
For example, when you register a method during development, S7 compares the method's formals to the generic's:

* The method's first arguments must match the generic's dispatch arguments, in order, with no defaults.
  A mismatch is an **error**.
* If the generic doesn't have `...`, the method's formals must match the generic's exactly.
  Any difference is an **error**.
* If the generic has a non-dispatch argument that the method lacks, that's a **warning**.
* If the method's default for an argument differs from the generic's, that's a **warning**.
* If the method has extra arguments that the generic lacks (and the generic has `...`), that's **silently allowed**.

But these checks only run during development, i.e. where someone can act on them:

* Under `devtools::load_all()`.
* Under `R CMD check`, but only when the package being checked owns the method, the generic, or one of the classes in the signature.
* When the method isn't in a package at all — registered in a script or interactively, as in this vignette's examples.

When you call `library(depPackage)` these checks are skipped, and the method is registered as-is.
This doesn't mean there's no problem, it just means that the problem is delayed enough for someone to do something about it.
This makes it possible to fix problems at run-time, not check-time.

## Signaling deprecation

S7 supplies three helpers for keeping an old interface available during a transition:

| Helper | With a replacement | Without a replacement |
|--------|--------------------|-----------------------|
| `deprecated_generic()` | Calls warn and dispatch through `new`; registrations target `new` silently. | Replace `new_generic()` in the existing definition, preserving its dispatch arguments and function. Calls warn and still dispatch. |
| `deprecated_class()` | Define the original class and supply `replacement` to recommend another class. Constructor calls warn and still create the original class. | Omit `replacement`; explicit constructor calls warn. |
| `deprecated_property()` | Reads and changed writes warn and use the property named by `new`. | Omit `new` to retain storage and warn on reads and changed writes. |

All three require `when`, the version of your package that introduced the deprecation, and accept the same `method` argument:

* `"base"`, the default, signals a `deprecatedWarning` on each deprecated call or access that triggers the helper.
* `"lifecycle(warn)"` uses `lifecycle::deprecate_warn()`, which throttles repeated warnings.
* `"lifecycle(stop)"` uses `lifecycle::deprecate_stop()` to make that use an error.

If you use a lifecycle method, add lifecycle to your package's dependencies.

Warnings do not by themselves make `R CMD check` fail, although they can cause tests to fail if they require silence or convert warnings to errors.
Use the warning phase to give downstream users a graceful transition period before switching to errors or removing an export.

## Changing a generic

### Changing dispatch

Changing a generic's dispatch arguments (i.e. their number, names, or order) breaks every downstream method.
There is no transition trick: a generic's dispatch arguments is its identity.
If you need to change dispatch, you will need to create a new generic then remove the old one, as shown later.

### Adding an argument

If your generic has `...` (and it should), adding a non-dispatch argument is one of the mildest changes you can make.
Existing methods keep working, but every registration produces a warning until they're updated:

```{r}
# myPackage, version 1
gen := new_generic("x")

# depPackage's method, written against version 1
BClass := new_class()
method(gen, BClass) <- function(x, ...) "old"

# myPackage, version 2 adds `verbose`
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) S7_dispatch())
method(gen, BClass) <- function(x, ...) "old"
```

Because the check only runs in development contexts, depPackage's users see nothing: installing and loading depPackage are both silent, and calls keep working.
The warning surfaces under `load_all()`, and as a single NOTE in `R CMD check` of depPackage:

```
* checking whether the namespace can be loaded with stated dependencies ... NOTE
Warning: gen(<depPackage::BClass>) doesn't have argument `verbose`
```

The fix is pleasingly asymmetric: because methods may have arguments the generic lacks, depPackage can add the argument **before** myPackage releases:

```{r}
# myPackage, version 1 again
gen := new_generic("x")

# depPackage adds `verbose` ahead of myPackage's release: no warning under version 1...
method(gen, BClass) <- function(x, verbose = FALSE, ...) "new"

# ...and none under version 2 either
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) S7_dispatch())
method(gen, BClass) <- function(x, verbose = FALSE, ...) "new"
```

So the transition recipe is:

1. You tell downstream authors about the upcoming argument (or send PRs).
2. Each downstream package adds the argument to its methods, and can release whenever convenient.
3. You release.

If your generic *doesn't* have `...`, adding an argument is an error for every downstream method, and there is no version of the method that satisfies both the old and the new generic.
Either way there is no fix on depPackage's side, so a generic without `...` is a frozen interface: only add one deliberately, and treat any change to it like a rename (below).

### Removing an argument

Removing a non-dispatch argument is silent at registration time: methods may have extra arguments, so a method that still lists the removed argument registers without complaint against both versions.
Downstream packages can remove the argument from their methods at leisure.

Callers are the bigger concern: `gen(x, removed_arg = 1)` will still be accepted by the generic (the argument passes through `...`) and then reach methods that may or may not expect it.
So before removing an argument, deprecate it: keep it in the generic, and warn when it's supplied:

```{r}
gen := new_generic("x", fun = function(x, verbose = FALSE, ...) {
  if (!missing(verbose)) {
    warning("`verbose` is deprecated.")
  }
  S7_dispatch()
})
method(gen, BClass) <- function(x, verbose = FALSE, ...) "result"
. <- gen(BClass(), verbose = TRUE)
```

If you use lifecycle, the equivalent is:

```{r}
gen := new_generic("x", fun = function(x, verbose = deprecated(), ...) {
  if (lifecycle::is_present(verbose)) {
    lifecycle::deprecate_warn("1.0.0", "gen(verbose)")
  }
  S7_dispatch()
})
method(gen, BClass) <- function(x, verbose = deprecated(), ...) "result"
. <- gen(BClass(), verbose = TRUE)
```

### Changing a default

Changing a default changes behavior immediately, for every caller, because S7 dispatch never uses the defaults provided by the method.
For this reason S7 warns whenever the generic and method drift out of sync:

```{r}
gen := new_generic("x", fun = function(x, drop = TRUE, ...) S7_dispatch())
method(gen, BClass) <- function(x, drop = FALSE, ...) drop
gen(BClass())
```

There's no way around this warning, as there's no spelling of the method that agrees with both versions of myPackage: whichever default it names disagrees with one of them, and omitting the default disagrees with both.
Fortunately the warning fires only in development contexts, which costs depPackage one check NOTE and costs depPackage's users nothing.

Our advice is to keep non-dispatch arguments to a minimum, and change defaults rarely, if ever.

### Renaming a generic

If you simply rename a generic, depPackage fails to install: its `importFrom(myPackage, gen)` directive (or `myPackage::gen` call) no longer resolves.

The fix is to keep the old name as a deprecated alias for the new generic:

```{r}
# myPackage, version 1
gen1 := new_generic("x")

# myPackage, version 2
gen2 := new_generic("x")
gen1 := deprecated_generic(new = gen2, when = "2.0.0")
method(gen1, BClass) <- function(x, ...) "result"
gen2(BClass())
gen1(BClass())
```

(If you plan to keep the old name indefinitely, a plain alias `gen1 <- gen2` is completely transparent and signals nothing.)

`deprecated_generic()` returns an object that still counts as the generic so method registration continues to work, but calls through the old name warn that it's time to move on.

### Removing a generic

To retire a generic without replacement, change `new_generic()` to `deprecated_generic()` and add `when`:

```{r}
retired := deprecated_generic("x", when = "2.0.0")
method(retired, BClass) <- function(x, ...) "result"
retired(BClass())
```

Existing method registration will continue to work (with a warning), giving dependencies time to move away from the generic before you remove in a future release.

### Moving a generic to another package

Splitting generics out into a lower-dependency package (say myPackageCore) works, as long as myPackage re-exports them:

```{r, eval = FALSE}
# myPackage, version 2
#' @importFrom myPackageCore gen
#' @export
myPackageCore::gen
```

depPackage's `importFrom(myPackage, gen)` still resolves, and method registration follows the generic object to its new home package automatically, 

Note that it's easy to write `gen <- myPackageCore::gen`, but this captures a copy of the generic when myPackage is installed. 
The result is a split brain: depPackage's registrations follow the generic's home package and land on myPackageCore's original, while calls through `myPackage::gen()` dispatch using the duplicate's empty table and fail.

Should the re-export signal deprecation?
Since a re-export costs nothing to maintain, keeping it forever is a fine choice.
If you do want to retire the old home, export a deprecated alias:

```{r, eval = FALSE}
# In myPackage, version 2
gen := deprecated_generic(new = myPackageCore::gen, when = "2.0.0")
```

## Changing a class

### Adding a property

Adding a property is usually straightforward. 
There are two ways it can still bite:

* **Name clash.** If depPackage's has a subclass and that subclass already has a property with the same name and its type doesn't narrow yours, depPackage will fail to install:

  ```
  Error: <MyFoo>@y must narrow <myPackage::Foo>@y.
  ```

  The easiest resolution to this problem is for you to pick a different name.
  If that doesn't work, you'll need to ask the depPackage maintainer to rename their property with the process described below.

* **Positional construction.** `myPackage::Foo(1, 2)` changes meaning if a property is inserted before an existing one.
  In general, it's safest to only add new properties at the end, but inserting might be ok if all callers used named arguments (as we recommend).

### Removing a property

Removing a property will cause run-time breakages: `Foo(count = 2)` will fail with `unused argument` and `x@count` with `Can't find property`.
To give downstream packages a transition window, deprecate the property with `deprecated_property()`:

```{r}
# myPackage, version 1
validate_count <- function(value) {
  if (any(value < 0)) "must be non-negative"
}
Foo := new_class(properties = list(
  count = new_property(class_double, default = 1, validator = validate_count)
))

# myPackage, version 2
Foo := new_class(properties = list(
  deprecated_property(
    "count", when = "2.0.0", class = class_double, default = 1,
    validator = validate_count
  )
))

foo <- Foo(count = 3)
foo@count
foo@count <- 4
print(foo)
```

### Renaming a property

A rename is "add the new property, deprecate the old one":

```{r}
Foo := new_class(properties = list(
  size = new_property(class_double, default = 1),
  deprecated_property("count", new = "size", when = "2.0.0")
))

foo <- Foo(size = 3) # silent
foo <- Foo(count = 3) # warns, then sets size
foo@count
foo@count <- 4
foo@size
```

The helper supplies both a getter and setter, so callers can keep using `Foo(count = ...)`, `foo@count`, and `foo@count <- ...` during the transition.
Put the replacement property first and preserve its old default and validation rules.
Put any `validator` on the replacement property.

### Changing a property's type or validator

Widening a property's type (say, `class_integer` to `class_numeric`) is safe for downstream code: everything that validated before still validates.
Narrowing is a run-time break: any code that stored a now-invalid value fails when it runs:

```
Error: <myPackage::Foo> object properties are invalid:
- @size must be <integer>, not <double>
```

The same applies to tightening a validator.
There's no mechanical transition here, but you can give downstream packages a deprecation window in which the validator warns rather than errors.

### Renaming a class

To rename, create a new class then change the previous definition to use `deprecated_class()`. Pass the replacement class definition to `replacement` and set the version `when` the replacement happened.

```{r}
# myPackage, version 1
Foo := new_class(properties = list(size = class_double))
saved <- Foo(size = 1)
MyFoo := new_class(parent = Foo)
measure := new_generic("x")
method(measure, Foo) <- function(x) x@size

# myPackage, version 2: export both classes
Bar := new_class(properties = list(size = class_double))
Foo := deprecated_class(
  properties = list(size = class_double),
  replacement = Bar,
  when = "2.0.0"
)
foo <- Foo(size = 2) # warns and still constructs a Foo
measure(foo)
measure(MyFoo(size = 3)) # existing subclasses still work, without warning
measure(saved)
S7_inherits(saved, Foo)
S7_inherits(saved, Bar)
```

Methods for `Foo` do not automatically apply to `Bar`.
When a method can handle both classes, register it on their union:

```{r}
method(measure, Foo | Bar) <- function(x) x@size
measure(Bar(size = 4))
measure(saved)
```

If the replacement needs a different representation, keep separate methods and provide an explicit conversion for users who want it.

### Moving a class

You can use the same technique to move a class between packages, because changing the class's `package` also changes its identity.
To recommend a class in another package while preserving existing uses, use `deprecated_class()`:

```{r, eval = FALSE}
# In myPackage
Foo := deprecated_class(
  properties = list(size = class_double),
  replacement = myPackageCore::Foo,
  when = "2.0.0"
)
```

### Removing a class

If you're removing a class without an alternative, omit `replacement` and keep the existing definition:

```{r}
Retired := deprecated_class(
  properties = list(size = class_double),
  when = "2.0.0"
)
Retired(size = 3)
```

Downstream subclasses, method signatures, property types, and unions can continue using the class but constructors will warn.

### Making a class abstract

Making an existing class abstract breaks every direct `myPackage::Foo()` call at run time, which is presumably the point of the change — give downstream packages a deprecation window by replacing `new_class()` with `deprecated_class()` in its definition, and only later make the class abstract.

Subclasses are unaffected, including already-installed ones: even though a subclass with an external parent builds its parent part by calling `myPackage::Foo()` at run time, an abstract class's constructor is allowed to run when it's constructing the parent part of a concrete subclass.
