---
title: "Using S7 in a package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using S7 in a package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette outlines the most important things you need to know about using S7 in a package.
S7 is new, so few people have used it in a package yet; this means that this vignette is likely incomplete, and we'd love your help to make it better.
Please [let us know](https://github.com/RConsortium/S7/issues/new) if you have questions that this vignette doesn't answer.

```{r setup}
library(S7)
```

## Method registration

You should always call `methods_register()` in your `.onLoad()`:

```{r}
.onLoad <- function(...) {
  S7::methods_register()
}
```

This is S7's way of registering methods, rather than using export directives in your `NAMESPACE` like S3 and S4 do.
This is only strictly necessary if registering methods for generics in other packages, but there's no harm in adding it and it ensures that you won't forget later.
(And if you're not importing S7 into your namespace it will quiet an `R CMD check` `NOTE`.`)`

## Documentation and exports

If you want users to create instances of your class, you will need to export the class constructor.
That means you will also need to document it, and since the constructor is a function, that means you have to document the arguments which will be the properties of the class (unless you have customised the constructor).

If you export a class, you must also set the `package` argument, ensuring that classes with the same name are disambiguated across packages.

You should document generics like regular functions (since they are!).
If you expect others to create their own methods for your generic, you may want to include an section describing the the properties that you expect all methods to have.
We plan to provide a an easy way to all methods for a generic, but have not yet implemented it.
You can track progress at <https://github.com/RConsortium/S7/issues/167>.

We don't currently have any recommendations on documenting methods.
There's no need to document them in order to pass `R CMD check`, but obviously there are cases where it's nice to provide additional details for a method, particularly if it takes extra arguments compared to the generic.
We're tracking that issue at <https://github.com/RConsortium/S7/issues/315>.

## Backward compatibility

If you are using S7 in a package *and* you want your package to work in versions of R before 4.3.0, you need to know that in these versions of R `@` only works with S4 objects.
There are two workarounds.
The easiest and least convenient workaround is to just `prop()` instead of `@`.
Otherwise, you can conditionally make an S7-aware `@` available to your package with this custom `NAMESPACE` directive:

``` r
# enable usage of <S7_object>@name in package code
#' @rawNamespace if (getRversion() < "4.3.0") importFrom("S7", "@")
NULL
```

`@` will work for users of your package because S7 automatically attaches an environment containing the needed definition when it's loaded.
