---
title: "Using S7 in a package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using S7 in a package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette outlines the most important things you need to know about using S7 in a package.
S7 is new, so few people have used it in a package yet; this vignette is likely incomplete, and we'd love your help to make it better.
Please [let us know](https://github.com/RConsortium/S7/issues/new) if you have questions that this vignette doesn't answer.

```{r setup}
library(S7)
```

## Getting started

First, add `S7` to the `Imports` field of your `DESCRIPTION`. We then recommend importing all S7 functions into your package `NAMESPACE` with `import(S7)`, or, if you're using roxygen2, `@import S7`.

Next, create a `zzz.R` with a `.onLoad()` that calls `S7_on_load()`, an `.onUnload()` that calls `S7_on_unload()`, and a top-level call to `S7_on_build()`:

```{r}
.onLoad <- function(...) {
  S7::S7_on_load()
}

.onUnload <- function(...) {
  S7::S7_on_unload()
}

S7::S7_on_build()
```

`S7_on_load()` is S7's way of registering methods, rather than using export directives in your `NAMESPACE` like S3 and S4 do.
This is only strictly necessary if you are registering methods for generics in other packages, but there's no harm in adding it and it ensures that you won't forget later.

`S7_on_unload()` cleans up after `S7_on_load()`: it unregisters methods that your package registered, if they are still active, and removes hooks, so unloading your package doesn't leave stale state behind.

`S7_on_build()` cleans up after `method<-`.
Because `method<-` is a replacement function, `method(generic, class) <- fun` is evaluated as `generic <- method<-(generic, class, fun)`, which would otherwise leave a copy of `generic` in your namespace.
When `generic` belongs to another package, `method<-` returns a sentinel value instead of the generic, and `S7_on_build()` removes these sentinels from your namespace when the package is built.
It must be called at the top level of `zzz.R` (i.e. *not* inside `.onLoad()`), after all your methods have been registered; since `zzz.R` is loaded last, this is usually automatic.

## Classes

If you want users to create instances of your class, you will need to export the class constructor.
That means you will also need to document it, and since the constructor is a function, you have to document the arguments, which will be the properties of the class (unless you have customized the constructor).
If you export a class (i.e. its constructor), you must also set the `package` argument, ensuring that classes with the same name are disambiguated across packages.

NB: if your package creates a hierarchy of classes, subclasses must be defined _after_ the parent classes. That means if you define the classes in separate files, you will need to use the `DESCRIPTION` Collate field (or the equivalent roxygen2 `@include` tag) to ensure the files are loaded in the correct order.

## Generics

You should document generics like regular functions (since they are!).
If you expect others to create their own methods for your generic, you may want to include a section describing the properties that you expect all methods to have.
If you want to list all methods for a generic, you can use the [doclisting](https://doclisting.r-lib.org) package.

## Methods

If you use roxygen2, you can document S7 generics and methods by following the advice in [`vignette("rd-S7", package = "roxygen2")`](https://roxygen2.r-lib.org/articles/rd-S7.html).

Note that methods can only be defined after both the class and generic have been defined. If generics/methods/classes live in different files, you will need to use the `DESCRIPTION` Collate field (or the equivalent roxygen2 `@include` tag) to ensure the files are loaded in the correct order.

### Methods for generics in suggested packages

If you want to register a method for a generic defined in a package that you only suggest (rather than import), use `new_external_generic()` to refer to the generic without taking a hard dependency:

```{r, eval = FALSE}
# In your package
median <- new_external_generic("stats", "median", "x")
method(median, MyClass) <- function(x, ...) { ... }
```

When the suggested package is loaded, S7 will register the method automatically (via `S7_on_load()` as described above).

### Methods for classes in suggested packages

Conversely, you may want to register a method for one of your own generics, dispatching on a class from a suggested package. Use `new_external_class()` to refer to the class by name:

```{r, eval = FALSE}
# In your package
my_generic <- new_generic("my_generic", "x")
ggplot <- new_external_class("ggplot2", "ggplot")
method(my_generic, ggplot) <- function(x) { ... }
```

When the suggested package is loaded, S7 will register the method automatically (via `S7_on_load()` as described above).

## Backward compatibility

### S3

S7 objects are S3 objects, so it's possible to manually register an S7 method for an S3 generic without using `method<-`. The main thing to note is that the S3 class name of an S7 object is `{package}::{class}`, which means you'll need to wrap the method name in `` ` ``:

```{r}
Foo := new_class(package = "S7")
`length.S7::Foo` <- function(x) 10
length(Foo())
```

### Older versions of R

If you are using S7 in a package *and* you want your package to work in versions of R before 4.3.0, you need to know that in these versions of R `@` only works with S4 objects.
There are two workarounds.
The easiest but least convenient workaround is to just use `prop()` instead of `@`.
Otherwise, you can conditionally make an S7-aware `@` available to your package with this custom `NAMESPACE` directive:

``` r
# enable usage of <S7_object>@name in package code
#' @rawNamespace if (getRversion() < "4.3.0") importFrom("S7", "@")
NULL
```

`@` will work for users of your package because S7 automatically attaches an environment containing the needed definition when it's loaded.
