---
title: "performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(R7)
```

The dispatch performance should be roughly on par with S3 and S4, though as this is implemented in a package there is some overhead due to `.Call` vs `.Primitive`.

```{r performance, cache = FALSE}
text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
number <- new_class("number", parent = "numeric", constructor = function(x) new_object(.data = x))

x <- text("hi")
y <- number(1)

foo_R7 <- new_generic(name = "foo_R7", signature = "x")
method(foo_R7, "text") <- function(x, ...) paste0(x, "-foo")

foo_s3 <- function(x, ...) {
  UseMethod("foo_s3")
}

foo_s3.text <- function(x, ...) {
  paste0(x, "-foo")
}

library(methods)
setOldClass(c("number", "numeric", "R7_object"))
setOldClass(c("text", "character", "R7_object"))

setGeneric("foo_s4", function(x, ...) standardGeneric("foo_s4"))
setMethod("foo_s4", c("text"), function(x, ...) paste0(x, "-foo"))

# Measure performance of single dispatch
bench::mark(foo_R7(x), foo_s3(x), foo_s4(x))

bar_R7 <- new_generic("bar_R7", c("x", "y"))
method(bar_R7, list("text", "number")) <- function(x, y, ...) paste0(x, "-", y, "-bar")

setGeneric("bar_s4", function(x, y, ...) standardGeneric("bar_s4"))
setMethod("bar_s4", c("text", "number"), function(x, y, ...) paste0(x, "-", y, "-bar"))

# Measure performance of double dispatch
bench::mark(bar_R7(x, y), bar_s4(x, y))
```

A potential optimization is caching based on the class names, but lookup should be fast without this.

The following benchmark generates a class hierarchy of different levels and lengths of class names and compares the time to dispatch on the first class in the hiearchy vs the time to dispatch on the last class.

We find that even in very extreme cases (e.g. 100 deep heirachy 100 of character class names) the overhead is reasonable, and for more reasonable cases (e.g. 10 deep hiearchy of 15 character class names) the overhead is basically negligible.

```{r performance-2, message = FALSE, R.options = list(width = 120), cache = TRUE}
library(R7)

gen_character <- function (n, min = 5, max = 25, values = c(letters, LETTERS, 0:9)) {
  lengths <- sample(min:max, replace = TRUE, size = n)
  values <- sample(values, sum(lengths), replace = TRUE)
  starts <- c(1, cumsum(lengths)[-n] + 1)
  ends <- cumsum(lengths)
  mapply(function(start, end) paste0(values[start:end], collapse=""), starts, ends)
}

bench::press(
  num_classes = c(3, 5, 10, 50, 100),
  class_nchar = c(15, 100),
  {
    # Construct a class hierarchy with that number of classes
    text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
    parent <- text
    classes <- gen_character(num_classes, min = class_nchar, max = class_nchar)
    env <- new.env()
    for (x in classes) {
      assign(x, new_class(x, parent = parent, constructor = function(text) new_object(.data = text)), env)
      parent <- get(x, env)
    }

    # Get the last defined class
    cls <- parent

    # Construct an object of that class
    x <- do.call(cls, list("hi"))

    # Define a generic and a method for the last class (best case scenario)
    foo_R7 <- new_generic(name = "foo_R7", signature = "x")
    method(foo_R7, cls) <- function(x, ...) paste0(x, "-foo")

    # Define a generic and a method for the first class (worst case scenario)
    foo2_R7 <- new_generic(name = "foo2_R7", signature = "x")
    method(foo2_R7, R7_object) <- function(x, ...) paste0(x, "-foo")

    bench::mark(
      best = foo_R7(x),
      worst = foo2_R7(x)
    )
  }
)
```

And the same benchmark using double-dispatch

```{r performance-3, message = FALSE, R.options = list(width = 120), cache = TRUE}
bench::press(
  num_classes = c(3, 5, 10, 50, 100),
  class_nchar = c(15, 100),
  {
    # Construct a class hierarchy with that number of classes
    text <- new_class("text", parent = "character", constructor = function(text) new_object(.data = text))
    parent <- text
    classes <- gen_character(num_classes, min = class_nchar, max = class_nchar)
    env <- new.env()
    for (x in classes) {
      assign(x, new_class(x, parent = parent, constructor = function(text) new_object(.data = text)), env)
      parent <- get(x, env)
    }

    # Get the last defined class
    cls <- parent

    # Construct an object of that class
    x <- do.call(cls, list("hi"))
    y <- do.call(cls, list("ho"))

    # Define a generic and a method for the last class (best case scenario)
    foo_R7 <- new_generic(name = "foo_R7", signature = c("x", "y"))
    method(foo_R7, list(cls, cls)) <- function(x, y, ...) paste0(x, y, "-foo")

    # Define a generic and a method for the first class (worst case scenario)
    foo2_R7 <- new_generic(name = "foo2_R7", signature = c("x", "y"))
    method(foo2_R7, list(R7_object, R7_object)) <- function(x, y, ...) paste0(x, y, "-foo")

    bench::mark(
      best = foo_R7(x, y),
      worst = foo2_R7(x, y)
    )
  }
)
```
