---
title: "Generalizing Support for Functional OOP in R"
output: html_document
editor_options:
  markdown:
    wrap: sentence
---

## Generalizing Support for Functional OOP in R

R has built-in support for two functional Object Oriented Programming (OOP) systems: S3 and S4, corresponding to the third and fourth version of the S language, respectively.
The two systems are largely compatible; however, they are two fundamentally distinct systems, two systems for a user to understand, two systems for a developer to contend with when making their package interoperate with another, two systems for R core to maintain.

[S7](https://rconsortium.github.io/S7/index.html) is a new OOP system being developed as a collaboration between representatives from R-Core, Bioconductor, tidyverse/Posit, ROpenSci, and the wider R community, with the goal of unifying S3 and S4 and promoting interoperability.
The long-term hope is to incorporate S7 into base R, following a period during which the concepts of S7 have been validated in a standalone R package.

The development of S7 has prompted its authors to reflect on how to generally enable functional OOP systems like S3, S4 and S7, as their remains a lot of room for innovation in this area.
Considering just the three mentioned systems, a number of commonalities become evident, including:

1.  Reifying classes, generics and methods as objects (in S4 and S7),

2.  Achieving polymorphism through generics dispatching to implementations (methods) based on one or more of their inputs, and

3.  Modeling objects as collections of components, such as slots or properties, that can be accessed and modified.

This blog post outlines four patches that have been integrated into base R to make it easier for an S3-based package like S7 to implement the above features.

Each of these four patches introduces a new S3 generic:

-   `chooseOpsMethod()`
-   `%*%`
-   `nameOfClass()`
-   `@`

This blog post describes these new generics, the features in S7 that motivated their implementation, and their utility to the broader R community outside of S7.
These enhancements are generally beneficial for packages integrating or implementing alternative OOP systems.
To illustrate these additional benefits, we use the reticulate package as an example.
Reticulate bridges Python's OOP semantics with R's S3 system.

### `chooseOpsMethod()`

S3 generics only support dispatching on their first argument, while S4 generics can dispatch on an arbitrary number of arguments.
S7 implements double dispatch (i.e. multiple dispatch on exactly two arguments) on top of S3 dispatch.
Dispatch in S7 is performed in two stages: S3 generics first dispatch on the first argument to an S7 aware method, which then invokes S7-internal dispatch routines that support multiple dispatch.
These are implementation details which, when everything works, are hidden from S7 users.
A general complication with implementing any sort of dispatch is that primitives and other functions with internal implementations must dispatch internally, in C code.
Internal dispatch and multiple dispatch intersect at the Ops group generics.

Unlike other S3 generics, the Ops group generics, consisting of infix binary operators like `+`, `-`, `*`, and so on, support dispatching to S3 methods on either the first or second argument.
Before R 4.3.0, if the two arguments would cause single dispatch to different methods, R would signal a warning and use neither.
Starting with R 4.3.0, if different methods are found for the first and second argument to the Ops generic, R will invoke `chooseOpsMethod()`, giving an opportunity for the object types to declare that their Ops method implementation can handle the combination.
If the `chooseOpsMethod()` method for an object returns `TRUE`, then the method found for that object is used by the Ops generic.

Because S7 implements its own generic dispatch mechanism, the `chooseOpsMethod()` method for an S7 object always returns `TRUE`.
This permits, for example, an S7 object type to define a `+` method that works with other S3 objects, like `Sys.Date()`.


```{r}
library(S7)

ClassX <- new_class("ClassX")

method(`+`, list(ClassX, class_any)) <- function(e1, e2) {
  "method: X + class_any"
}

x <- ClassX()

x + Sys.Date()
```

This also has utility for other packages that implement alternative OO systems.
For example, reticulate is an R package that embeds Python in an R session, and allows R users to operate with Python objects directly.
Python has its own implementation of method dispatch for infix operators like `+`.
The addition of `chooseOpsMethod()` enables reticulate to implement a default fallback `+` method that dispatches to the corresponding Python routines, while still allowing R users to define a custom `+` method for specific Python object types.

For example, the `+` operator can now select the appropriate method when passed a TensorFlow Tensor and some other Python object.
Previously, dispatch failed, because both arguments are instances of an S3 class but would cause dispatch to different methods.


```{r}
library(reticulate)

# Adding a NumPy array and an R array dispatches to the default
# `+` method for "python.builtin.object"
np_array(1:3) + array(1:3)

# Adding a TensorFlow Tensor and an R array invokes
# invokes the specific `+` method for "tensorflow.tensor"
array(1:3) + tensorflow::as_tensor(1:3)

# Adding a NumPy array with a TensorFlow Tensor.
# Prior to R 4.3.0, this would signal an error and warn:
#   Incompatible methods ("+.tensorflow.tensor", "+.python.builtin.object")
# Beginning with R 4.3.0, chooseOpsMethod() is called to choose the appropriate
# Ops method, and this now works
np_array(1:3) + tensorflow::as_tensor(1:3)
```

### `%*%`

Base R defines additional operators that dispatch internally and support some form of multiple dispatch, but are not in the Ops group. The matrix multiplication operator, `%*%`, is a prime example.

Prior to R 4.3.0, `%*%` only worked with S4 objects or base R objects.
Now, S3 methods for this operator can be defined as well.
Just like the Ops group generics, `%*%` will dispatch on the second argument, and use `chooseOpsMethod()` to resolve conflicts.

To enable this new behavior, a new generics group was created: `matrixOps`.
Presently, the sole member of the group is `%*%`, though in the future, `crossprod()` and `tcrossprod()` are expected to join the group.

The addition of the `"matrixOps"` group is a narrow backwards-compatible change, intended to enable existing operators to work with S3 and S7.
Future functions that aim to support double dispatch via S3 will be able to make use of this general mechanism, rather than requiring further changes to R's S3 semantics.

### `nameOfClass()`

Though S7 uses S3 dispatch, S7 reifies classes as objects, and its user interface is based on those objects, instead of class names.
The class name stored in the class vector for S3 dispatch is considered an implementation detail (it is non-syntatic and mangled to minimize the potential for conflicts, and not intended to be user-facing).
Additionally, it can be subject to change if the generic definition migrates between packages.

For these reasons, it is desirable to be able to check for inheritance without directly using the S3 class name.
To enable this for S7 and potentially other packages with their own class representations, beginning with R 4.3.0 `base::inherits()` now accepts arbitrary objects for the second argument: if the second argument is an S3 object, the new `nameOfClass()` S3 generic is invoked to resolve the appropriate class name.

This permits usage in S7 like this:


```{r}
ClassX <- S7::new_class("ClassX")

x <- ClassX()

inherits(x, ClassX)
```

This also has utility for other R packages, in particular, packages where an external class hierarchy is being mirrored in R.
For example, in reticulate, the S3 class vector for Python object is generated from the Python module where the class is defined.
However, in sophisticated codebases like Keras and TensorFlow, the location where a class is defined in the Python sources is considered an internal implementation detail, and often changes between library versions (i.e., the module where a class is defined is decoupled from the module where users are meant to access the class).
To ensure compatibility between and across Python library versions, reticulate now implements a `nameOfClass()` method which enables R usage like this:


```{r}
library(tensorflow)
x <- tf$Variable(1:3)

inherits(x, tf$Variable)
```

### `@`

Prior to R 4.3.0, `@` could be used to access slots of S4 objects.
Beginning with R 4.3.0, this operator can now perform S3 dispatch as well.
This bring `@` to parity with `@<-` (which has long been able to do S3 dispatch), and makes `@` a peer (in dispatch-ability) of `$`.
This is a strictly backwards compatible change, as S3 dispatch is only performed *after* S4 routines have had an opportunity to dispatch.

The ability to have formally defined slots or properties is one of the benefits of S4 and S7; it enables collaborative work at a greater scale than the ad-hoc conventions of S3 classes, and having a dedicated operator for property access in S7 is desirable.
We expect other OOP systems will have similar concepts and also benefit from this generalization.
Other than performing S3 dispatch, there are no other user-facing changes with `@` in base R---the default behavior of `@` in the absence of an S3 method is unchanged.
