---
title: "Introducing R7"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing R7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The R7 package is a new OOP system designed to be a successor to S3 and S4.
It has been designed and implemented collaboratively by the RConsortium Object-Oriented Programming Working Group, which includes representatives from R-Core, BioConductor, RStudio/tidyverse, and the wider R community.

This vignette describes how R7 works, starting with an overview of classes, objects, generics, and methods, before going into the details of validation and properties.

```{r setup}
library(R7)
```

## Basics

We'll start with an quick overview of the most important parts of R7: classes and objects, generics and methods, and the basics of method dispatch and inheritance.

### Classes and objects

R7 classes have a formal definition that you create with `new_class()`.
There are two arguments that you'll use with almost every class:

-   The `name`, of the class, supplied in the first argument.
-   The class `properties`, which is the data associated with each instance of each instance of the class. The easiest way to define properties is to supply a named list where the values define the valid types of the property.

The following code defines a simple `pet` class with two properties: a character `name` and a double `age`.

```{r}
dog <- new_class("dog", properties = list(
  name = character,
  age = double
))
dog
```

Note that I've assigned the return value of `new_class()` to an object with the same name as the class.
This is important!
That object represents the class and is what you use to construct instances of the class:

```{r}
lola <- dog(name = "Lola", age = 11)
lola
```

Once you have an R7 object, you can get and set properties using `@`:

```{r}
lola@age <- 11
lola@age
```

R7 automatically validates the type of the property using the type supplied in `new_class()`:

```{r, error = TRUE}
lola@age <- "twelve"
```

You can retrieve the class that the object is an instance of with `R7_class()`:

```{r}
R7_class(lola)
```

### Generics and methods

R7, like S3 and S4, is built around the idea of a **generic function,** or **generic** for short.
A generic defines an interface, which uses a different implementation, depending on the class of an argument (usually the first argument).
The implementation for a specific class is called a **method**, and the generic finds that method by performing **method dispatch**.

Use `new_generic()` to create a R7 generic.
In its simplest form, it only needs two arguments: the name of the generic (used in error messages) and the name of the argument it should dispatch on:

```{r}
speak <- new_generic("speak", "x")
```

Like with `new_class()`, you should always assign the result of `new_generic()` to an variable with the same name as the first argument.

Once you have a generic, you can register methods for specific classes with `method<-`.
The first argument to `method<-` is the generic, and the second is the class:

```{r}
method(speak, dog) <- function(x) {
  "Woof"
}
```

Once registered, you can call the generic with an instance of that class to use the method:

```{r}
speak(lola)
```

Let's define another class, this one for cats, and define another method:

```{r}
cat <- new_class("cat", properties = list(
  name = character,
  age = double
))
method(speak, cat) <- function(x) {
  "Meow"
}

fluffy <- cat(name = "Fluffy", age = 5)
speak(fluffy)
```

If you call the generic with a class that isn't registered, you'll get an error:

```{r, error = TRUE}
speak(1)
```

### Method dispatch and inheritance

The `cat` and `dog` classes share the same properties, so we could use a common parent class to extract out the duplicated specification.
Now we use the second argument to `new_class` to specify the `parent` class:

```{r}
pet <- new_class("pet", properties = list(
  name = character,
  age = double
)) 
cat <- new_class("cat", pet)
dog <- new_class("dog", pet)

cat
dog
```

Because we have created new classes, we need to recreate the existing `lola` and `fluffy` objects:

```{r}
lola <- dog(name = "Lola", age = 11)
fluffy <- cat(name = "Fluffy", age = 5)
```

Method dispatch takes advantage of the hierarchy of parent classes: if a method is not defined for a class, it will try the method for the parent class, and so on until it either finds a method or gives up.
This inheritance makes it possible to share code across classes.

```{r}
describe <- new_generic("describe", "x")
method(describe, pet) <- function(x) {
  paste0(x@name, " is ", x@age, " years old")
}
describe(lola)
describe(fluffy)

method(describe, dog) <- function(x) {
  paste0(x@name, " is a ", x@age, " year old dog")
}

describe(lola)
describe(fluffy)
```

You can define a fallback for any R7 class by implementing a method for `R7_object`:

```{r}
method(describe, R7_object) <- function(x) {
  "An R7 object"
}

cocktail <- new_class("cocktail", 
  properties = list(
    ingredients = character
  )
)
martini <- cocktail(ingredients = c("gin", "vermouth"))
describe(martini)
```

Printing a generic will show you which methods are currently defined:

```{r}
describe
```

And you can use `method()` to retrieve the implementation of one of those methods:

```{r}
method(describe, pet)
```

Learn more about method dispatch in `vignette("dispatch")`.

## Validation

R7 classes can have an optional **validator** that checks that the values of the properties are ok.
A validator is a function that takes an object (called `self`) and returns `NULL` if the object is valid, or if not, returns a character vector listing the problems.

In the following example we create a range class that enforces that the start and end are single numbers where start is less than end:

```{r}
range <- new_class("range",
  properties = list(
    start = "numeric",
    end = "numeric"
  ),
  validator = function(self) {
    if (length(self@start) != 1) {
      "@start must be length 1"
    } else if (length(self@end) != 1) {
      "@end must be length 1"
    } else if (self@end < self@start) {
      sprintf(
        "@end (%i) must be greater than or equal to @start (%i)",
        self@end,
        self@start
      )
    }
  }
)
```

You can typically write a validator as a series of `if`-`else` statements, but note that the order of the statements is important.
For example, in the code above, we can't check that `self@end < self@start` before we've checked that `@start` and `@end` are length 1.

Objects are validated automatically when constructed and when any property is modified:

```{r, error = TRUE}
x <- range(1, 2:3)
x <- range(10, 1)

x <- range(1, 10)
x@start <- 20
```

You can manually `validate()` an object if you use a low-level R function to bypass the usual checks and balances of `@`:

```{r, error = TRUE}
x <- range(1, 2)
attr(x, "start") <- 3
validate(x)
```

Imagine you wanted to write a function that would shift a property to the left or the right:

```{r}
shift <- function(x, shift) {
  x@start <- x@start + shift
  x@end <- x@end + shift
  x
}
shift(range(1, 10), 1)
```

There's a problem if `shift` is larger than `@end` - `@start`:

```{r, error = TRUE}
shift(range(1, 10), 10)
```

While the end result of `shift()` will be valid, an intermediate state is not.
The easiest way to resolve this problem is to set the properties all at once:

```{r}
shift <- function(x, shift) {
  props(x) <- list(
    start = x@start + shift,
    end = x@end + shift
  )
  x
}
shift(range(1, 10), 10)
```

The object is still validated, but it's only validated once, after all the properties have been modified.

## Properties

So far we've focused on the simplest form of property specification where you use a named list to supply the desired type for each property.
This is shorthand for a call to `new_property()`.
For example, the property definition of range above is shorthand for:

```{r}
range <- new_class("range",
  properties = list(
    start = new_property("numeric"),
    end = new_property("numeric")
  )
)
```

Using `new_property()` allows you to control aspects of the property apart from its type.
The following sections show you to provide a default value, compute the property value on demand, and provide a fully dynamic property.

### Types

-   Avoid any restriction with with `any_class`.
-   Refer to base types using their constructors: `logical`, `integer`, `double`, `character` etc.
-   Use a string to refer to union base types with `"numeric"` (`integer` or `double`), `"atomic"` (`logical`, `"numeric"`, or `character`), or `"vector"` (`"atomic"` or `list`).
-   Use `new_union()` to create your own set of possible values like `new_union("integer", "NULL")`.
-   Require an S3 class with `new_S3_class("class_name")`.
-   Require an S4 class by providing it a class definition or generator.

See `vignette("compatibility")` for more details about using S3 and S4 with R7.

### Default value

The defaults of `new_class()` create an class that can be constructed with no arguments:

```{r}
empty <- new_class("empty", 
  properties = list(
    x = double,
    y = character,
    z = logical
  ))
empty()
```

The default values of the properties will be filled in with empty instances of their types.
You can provide your own default by using the `default` argument to `new_property()`:

```{r}
empty <- new_class("empty", 
  properties = list(
    x = new_property(numeric, default = 0),
    y = new_property(character, default = ""),
    z = new_property(logical, default = NA)
  )
)
empty()
```

### Computed properties

It's sometimes useful to have a property that is computed on demand.
For example, it'd be convenient to pretend that our range has a length, which is just the distance between `@start` and `@end`.
You can dynamically compute the value of a property by defining a `getter`:

```{r}
range <- new_class("range",
  properties = list(
    start = "numeric",
    end = "numeric",
    length = new_property(
      getter = function(self) self@end - self@start,
    )
  )
)

x <- range(start = 1, end = 10)
x
```

Computed property are is read-only:

```{r, error = TRUE}
x@length <- 20
```

### Dynamic properties

You can make a computed property fully dynamic so that it can be read and written by supplying a `setter`.
For example, we could extend the previous example to allow the `@length` to be set, by modifying the `@end` of the vector:

```{r}
range <- new_class("range",
  properties = list(
    start = "numeric",
    end = "numeric",
    length = new_property(
      class = "numeric",
      getter = function(self) self@end - self@start,
      setter = function(self, value) {
        self@end <- self@start + value
        self
      }
    )
  )
)

x <- range(start = 1, end = 10)
x

x@length <- 5
x
```

A `setter` is a function with arguments `self` and `value` that returns a modified object.
