---
title: "S7 basics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S7 basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The S7 package provides a new OOP system designed to be a successor to S3 and S4.
It has been designed and implemented collaboratively by the RConsortium Object-Oriented Programming Working Group, which includes representatives from R-Core, BioConductor, RStudio/tidyverse, and the wider R community.

This vignette describes how S7 works, starting with an overview of classes, objects, generics, and methods, then going into the details of validation and properties.

```{r setup}
library(S7)
```

## Basics

We'll start with an quick overview of the most important parts of S7: classes and objects, generics and methods, and the basics of method dispatch and inheritance.

### Classes and objects

S7 classes have a formal definition that you create with `new_class()`.
There are two arguments that you'll use with almost every class:

-   The `name` of the class, supplied in the first argument.
-   The class `properties`, the data associated with each instance of the class. The easiest way to define properties is to supply a named list where the values define the valid types of the property.

The following code defines a simple `dog` class with two properties: a character `name` and a numeric `age`.

```{r}
dog <- new_class("dog", properties = list(
  name = class_character,
  age = class_numeric
))
dog
```

S7 provides a number of built-in definitions that allow you to refer to existing base types that are not S7 classes.
You can recognize these definitions because they all start with `class_`.

Note that I've assigned the return value of `new_class()` to an object with the same name as the class.
This is important!
That object represents the class and is what you use to construct instances of the class:

```{r}
lola <- dog(name = "Lola", age = 11)
lola
```

Once you have an S7 object, you can get and set properties using `@`:

```{r}
lola@age <- 12
lola@age
```

S7 automatically validates the type of the property using the type supplied in `new_class()`:

```{r, error = TRUE}
lola@age <- "twelve"
```

Given an object, you can retrieves its class `S7_class()`:

```{r}
S7_class(lola)
```

S7 objects also have an S3 `class()`.
This is used for compatibility with existing S3 generics and you can learn more about it in `vignette("compatibility")`.

```{r}
class(lola)
```

### Generics and methods

S7, like S3 and S4, is built around the idea of **generic functions,** or **generics** for short.
A generic defines an interface, which uses a different implementation depending on the class of one or more arguments.
The implementation for a specific class is called a **method**, and the generic finds that appropriate method by performing **method dispatch**.

Use `new_generic()` to create a S7 generic.
In its simplest form, it only needs two arguments: the name of the generic (used in error messages) and the name of the argument used for method dispatch:

```{r}
speak <- new_generic("speak", "x")
```

Like with `new_class()`, you should always assign the result of `new_generic()` to a variable with the same name as the first argument.

Once you have a generic, you can register methods for specific classes with `method(generic, class) <- implementation`.

```{r}
method(speak, dog) <- function(x) {
  "Woof"
}
```

Once the method is registered, the generic will use it when appropriate:

```{r}
speak(lola)
```

Let's define another class, this one for cats, and define another method for `speak()`:

```{r}
cat <- new_class("cat", properties = list(
  name = class_character,
  age = class_double
))
method(speak, cat) <- function(x) {
  "Meow"
}

fluffy <- cat(name = "Fluffy", age = 5)
speak(fluffy)
```

You get an error if you call the generic with a class that doesn't have a method:

```{r, error = TRUE}
speak(1)
```

### Method dispatch and inheritance

The `cat` and `dog` classes share the same properties, so we could use a common parent class to extract out the duplicated specification.
We first define the parent class:

```{r}
pet <- new_class("pet",
  properties = list(
    name = class_character,
    age = class_numeric
  )
)
```

Then use the `parent` argument to `new_class:`

```{r}
cat <- new_class("cat", parent = pet)
dog <- new_class("dog", parent = pet)

cat
dog
```

Because we have created new classes, we need to recreate the existing `lola` and `fluffy` objects:

```{r}
lola <- dog(name = "Lola", age = 11)
fluffy <- cat(name = "Fluffy", age = 5)
```

Method dispatch takes advantage of the hierarchy of parent classes: if a method is not defined for a class, it will try the method for the parent class, and so on until it finds a method or gives up with an error.
This inheritance is a powerful mechanism for sharing code across classes.

```{r}
describe <- new_generic("describe", "x")
method(describe, pet) <- function(x) {
  paste0(x@name, " is ", x@age, " years old")
}
describe(lola)
describe(fluffy)

method(describe, dog) <- function(x) {
  paste0(x@name, " is a ", x@age, " year old dog")
}
describe(lola)
describe(fluffy)
```

You can define a fallback method for any S7 object by registering a method for `S7_object`:

```{r}
method(describe, S7_object) <- function(x) {
  "An S7 object"
}

cocktail <- new_class("cocktail",
  properties = list(
    ingredients = class_character
  )
)
martini <- cocktail(ingredients = c("gin", "vermouth"))
describe(martini)
```

Printing a generic will show you which methods are currently defined:

```{r}
describe
```

And you can use `method()` to retrieve the implementation of one of those methods:

```{r}
method(describe, pet)
```

Learn more about method dispatch in `vignette("dispatch")`.

## Validation

S7 classes can have an optional **validator** that checks that the values of the properties are OK.
A validator is a function that takes the object (called `self`) and returns `NULL` if its valid or returns a character vector listing the problems.

### Basics

In the following example we create a range class that enforces that `@start` and `@end` are single numbers, and that `@start` is less than `@end`:

```{r}
range <- new_class("range",
  properties = list(
    start = class_double,
    end = class_double
  ),
  validator = function(self) {
    if (length(self@start) != 1) {
      "@start must be length 1"
    } else if (length(self@end) != 1) {
      "@end must be length 1"
    } else if (self@end < self@start) {
      sprintf(
        "@end (%i) must be greater than or equal to @start (%i)",
        self@end,
        self@start
      )
    }
  }
)
```

You can typically write a validator as a series of `if`-`else` statements, but note that the order of the statements is important.
For example, in the code above, we can't check that `self@end < self@start` before we've checked that `@start` and `@end` are length 1.

### When is validation performed?

Objects are validated automatically when constructed and when any property is modified:

```{r, error = TRUE}
x <- range(1, 2:3)
x <- range(10, 1)

x <- range(1, 10)
x@start <- 20
```

You can also manually `validate()` an object if you use a low-level R function to bypass the usual checks and balances of `@`:

```{r, error = TRUE}
x <- range(1, 2)
attr(x, "start") <- 3
validate(x)
```

### Avoiding validation

Imagine you wanted to write a function that would shift a property to the left or the right:

```{r}
shift <- function(x, shift) {
  x@start <- x@start + shift
  x@end <- x@end + shift
  x
}
shift(range(1, 10), 1)
```

There's a problem if `shift` is larger than `@end` - `@start`:

```{r, error = TRUE}
shift(range(1, 10), 10)
```

While the end result of `shift()` will be valid, an intermediate state is not.
The easiest way to resolve this problem is to set the properties all at once:

```{r}
shift <- function(x, shift) {
  props(x) <- list(
    start = x@start + shift,
    end = x@end + shift
  )
  x
}
shift(range(1, 10), 10)
```

The object is still validated, but it's only validated once, after all the properties have been modified.

## Properties

So far we've focused on the simplest form of property specification where you use a named list to supply the desired type for each property.
This is a convenient shorthand for a call to `new_property()`.
For example, the property definition of range above is shorthand for:

```{r}
range <- new_class("range",
  properties = list(
    start = new_property(class_double),
    end = new_property(class_double)
  )
)
```

Calling `new_property()` explicitly allows you to control aspects of the property other than its type.
The following sections show you how to provide a default value, compute the property value on demand, or provide a fully dynamic property.

### Default value

The defaults of `new_class()` create an class that can be constructed with no arguments:

```{r}
empty <- new_class("empty",
  properties = list(
    x = class_double,
    y = class_character,
    z = class_logical
  ))
empty()
```

The default values of the properties will be filled in with "empty" instances.
You can instead provide your own defaults by using the `default` argument:

```{r}
empty <- new_class("empty",
  properties = list(
    x = new_property(class_numeric, default = 0),
    y = new_property(class_character, default = ""),
    z = new_property(class_logical, default = NA)
  )
)
empty()
```

### Computed properties

It's sometimes useful to have a property that is computed on demand.
For example, it'd be convenient to pretend that our range has a length, which is just the distance between `@start` and `@end`.
You can dynamically compute the value of a property by defining a `getter`:

```{r}
range <- new_class("range",
  properties = list(
    start = class_double,
    end = class_double,
    length = new_property(
      getter = function(self) self@end - self@start,
    )
  )
)

x <- range(start = 1, end = 10)
x
```

Computed properties are read-only:

```{r, error = TRUE}
x@length <- 20
```

### Dynamic properties

You can make a computed property fully dynamic so that it can be read and written by also supplying a `setter`.
For example, we could extend the previous example to allow the `@length` to be set, by modifying the `@end` of the vector:

```{r}
range <- new_class("range",
  properties = list(
    start = class_double,
    end = class_double,
    length = new_property(
      class = class_double,
      getter = function(self) self@end - self@start,
      setter = function(self, value) {
        self@end <- self@start + value
        self
      }
    )
  )
)

x <- range(start = 1, end = 10)
x

x@length <- 5
x
```

A `setter` is a function with arguments `self` and `value` that returns a modified object.
