---
title: "Dispatch"
description: Initial notes on dispatch design
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

## Single dispatch

### S3

The basic rules of S3 dispatch are simple.
If object has class attribute `c("a", "b", "c")` then generic `f()` looks for methods in the following order:

-   `f.a()`
-   `f.b()`
-   `f.c()`
-   `f.default()`

If no method is found, it errors.

### S7

S7 will behave the same as S3.

## Method lookup

### S3

Where precisely does `UseMethod()` look for the methods?
As of R 4.0.0, it looks in the following three places:

-   The method table is a special environment `.__S3MethodsTable__.` found in the environment where the generic is defined.

-   The chain of environments starting at the `parent.frame()` of the call to generic, ending at the global environment.

-   The base environment (i..e. skips the search).

### S7

S7 methods are defined using assignment:

```{r, eval = FALSE}
method("mean", "numeric") <- function(x) sum(x) / length(x)
```

Behind the scenes, this acts directly upon the method table, so method lookup for S7 generics never needs to look in the parent frame.

`method<-` is likely to start as a shim around `.S3method()` but we may want to consider a separate `.__S7MethodsTable__.`.
This could use a new data structure that resolves generic/class ambiguity (e.g. `all.equal.data.frame()`).
Methods for S7 classes defined on an S3 generics would still use the S3 method table.
Could consider attaching the method table to the generic, instead of its containing environment.

Method lookup would be cached for performance, so that it is only performed once per class.
Cached methods would be marked with a special attribute so that they could be flushed whenever a new method for the generic is added.

## Method call frame

### S3

Once the method has been found, it must be called.
`UseMethod()` does not work like a regular function call but instead:

-   Changes to arguments are ignored.

-   Method can access objects created in generic.

-   The parent frame of the method call is the parent frame of the generic.

These properties are summarised in the following example:

```{r}
foo <- function(x, y)  {
  y <- 2
  z <- 2
  UseMethod("foo")
}
foo.numeric <- function(x, y) {
  print(parent.frame())
  c(x = x, y = y, z = z)
}
foo(1, 1)
```

### S7

-   Can we eliminate the special behaviour and make it just like a regular function call?
    Presumably easier than changing dispatch rules because we'll call a function other than `UseMethod()`.

-   Need to make precise how arguments are passed to the method.
    `plot()` at least assumes that this works:

    ```{r}
    foo <- function(x, y)  {
      UseMethod("foo")
    }
    foo.numeric <- function(x, y) {
      deparse(substitute(x))
    }
    x <- 10
    foo(x)
    ```

    How does that intersect with assignment within the generic?

## Inheritance

### S3

i.e. how does `NextMethod()` work: currently most state recorded in special variables like `.Generic`, etc.

Can we avoid this confusion:

```{r}
foo <- function(x)  {
  UseMethod("foo")
}
foo.a <- function(x) {
  x <- factor("x")
  NextMethod()
}
foo.b <- function(x) {
  print("In B")
  print(class(x))
}

foo(structure(1, class = c("a", "b")))
```

### S4

Want to avoid this sort of code, where we rely on magic from `callGeneric()` to pass on values from current call.

```{r, eval = FALSE}
method("mean", "foofy") <- function(x, ..., na.rm = TRUE) {
  x <- x@values
  callGeneric()
}
```

### S7

Can we require `generic` and `object` arguments to make code easier to reason about?

```{r, eval = FALSE}
method("mean", "POSIXct") <- function(x) {
  POSIXct(NextMethod(), tz = attr(x, "tz"))
}
# Explicit is nice:
method("mean", "POSIXct") <- function(x) {
  POSIXct(NextMethod("mean", x), tz = attr(x, "tz"))
}
# But what does this do? Is this just an error?
method("mean", "POSIXct") <- function(x) {
  POSIXct(NextMethod("sd", 10), tz = attr(x, "tz"))
}
```

## Group generics

### S3

Group generics (`Math`, `Ops`, `Summary`, `Complex`): exist for some internal generics.
Looked for before final fallback.

```{r}
sloop::s3_dispatch(sum(Sys.time()))
```

### S7

Keep as is.

## Double dispatch

### S3

Used by Ops group generic.
Basic process is find method for first and second arguments.
Then:

-   If same, ok
-   If one internal, use other
-   Otherwise, warn and use internal

### S7

Goal is to use iterated dispatch which implies asymmetry in dispatch order.
User responsible for ensuring that `x + y` equivalent to `y + x` (types should almost always be the same, but values are likely to be different).

```{r}
double_dispatch <- function(x, y, generic = "+") {
  grid <- rev(expand.grid(sloop::s3_class(y), sloop::s3_class(x)))
  writeLines(paste0("* ", generic, ".", grid[[1]], ".", grid[[2]]))
}

ab <- structure(list(), class = c("a", "b"))
cd <- structure(list(), class = c("c", "d"))

double_dispatch(ab, cd)
double_dispatch(cd, ab)

double_dispatch(1, 1L)
```

In vctrs, some question if we will remove inheritance from all double dispatch.
We have already done so for `vec_ptype2()` and `vec_cast()` because the coercion hierarchy often does not match the class hierarchy.
May also do for `vec_arith()`.

## Implicit class

### S3

When `UseMethod()` receives an object without a `class` attribute, it uses the **implicit** class, as provided by `.class2()`.
This is made up of four rough categories: dimension, type, language, numeric.

```{r}
# dimension class
.class2(matrix("a"))
.class2(array("a"))

# typeof(), with some renaming
.class2(sum)
.class2(quote(x))

# language class
.class2(quote({}))
# similarly for if, while, for, =, <-, (

# numeric
.class2(1)
```

Note that internal generics behave differently, instead immediately falling back to the default default case.

### S7

Suggest defining a new `r7class()` function that returns a simplified implicit class, dropping the language classes.

Dispatch should use the same rules in R and in C.
(But are there performance implications?)

## Multi-dispatch

### S3

Special dispatch?
`c()`, `cbind()`, `rbind()` (+ `cbind2()` and `rbind2()`) --- iterated double dispatch.
Need to describe in more detail so we have a more solid assessment of what S7 might need.ez

-   gitDot-dot-dot dispatch, assumes all have same class

-   vctrs used two pass approach (find type then coerce)

### S7

Initially, don't provide support for user generics that dispatch on `…`?
Instead suggest people use `Reduce` plus double-dispatch.
